---
title: "fastder-evaluation runtime and memory"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  bench_dir: NULL
  scenarios: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 6,
  fig.align = "center"
)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)

pub_theme <- theme_bw(base_size = 12) +
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_line(colour = "grey88"),
    panel.grid.minor = element_line(colour = "grey94"),
    strip.background = element_rect(fill = "grey95", colour = NA),
    plot.title = element_text(size = 13, face = "plain"),
    legend.position = "none"
  )

TOOLS <- c("fastder", "derfinder", "megadepth_baseline", "grohmm")

# Tool palette. Same hex values as summary.Rmd so the two reports stay
# consistent. Keys are the tool names as they appear in the benchmark TSVs.
tool_palette <- c(
  derfinder          = "#66C2A5",
  fastder            = "#FC8D62",
  grohmm             = "#8DA0CB",
  megadepth_baseline = "#E78AC3"
)
```

```{r load}
bench_dir <- params$bench_dir
if (is.null(bench_dir) || !dir.exists(bench_dir)) {
  stop("bench_dir parameter must point to an existing benchmarks directory")
}
files <- list.files(bench_dir, pattern = "\\.tsv$", recursive = TRUE, full.names = TRUE)
if (length(files) == 0) {
  stop("no .tsv benchmark files found under ", bench_dir)
}
parse_one <- function(path) {
  rel <- sub(paste0(bench_dir, .Platform$file.sep), "", path)
  parts <- strsplit(rel, .Platform$file.sep, fixed = TRUE)[[1]]
  if (length(parts) == 1) {
    rule <- sub("\\.tsv$", "", parts[1])
    wc   <- ""
  } else {
    rule <- parts[1]
    wc   <- sub("\\.tsv$", "", paste(parts[-1], collapse = "_"))
  }
  d <- read_tsv(path, show_col_types = FALSE)
  d$rule <- rule
  d$wildcards <- wc
  d
}
df <- bind_rows(lapply(files, parse_one))
df$wall_s    <- as.numeric(df$s)
df$max_rss_mb <- as.numeric(df$max_rss)

# Derive a `tool` column for rules whose benchmark files encode the tool.
# - run_fastder, run_derfinder, run_megadepth_baseline, run_grohmm: tool =
#   rule suffix.
# - run_gffcompare, eval_fuzzy_metrics, link_fastder_gtf: wildcards begin
#   with the tool name, so we extract the leading token.
tool_from_rule <- function(rule) {
  out <- rep(NA_character_, length(rule))
  out[rule == "run_fastder"] <- "fastder"
  out[rule == "run_derfinder"] <- "derfinder"
  out[rule == "run_megadepth_baseline"] <- "megadepth_baseline"
  out[rule == "run_grohmm"] <- "grohmm"
  out
}
tool_from_wildcards <- function(wc) {
  m <- str_match(wc, paste0("^(", paste(TOOLS, collapse = "|"), ")_"))
  m[, 2]
}
df$tool <- tool_from_rule(df$rule)
needs_wc_tool <- df$rule %in% c("run_gffcompare", "eval_fuzzy_metrics", "link_fastder_gtf")
df$tool[needs_wc_tool] <- tool_from_wildcards(df$wildcards[needs_wc_tool])

# Drop the evaluation-harness rules from the benchmark displays. They are
# not methods under test, just the gffcompare/fuzzy-eval scaffolding the
# pipeline always runs once per (tool, scenario, sample, param_id).
df <- df %>% filter(!rule %in% c("run_gffcompare", "eval_fuzzy_metrics"))

# Scenario names for this run, supplied by the rule from the config. Match
# them generally so the report works for any config (simulation scenarios,
# gtex tissue sub-groups, recount3 groups), not a hardcoded pair. Longest
# first so a scenario that prefixes another (brain_1 vs brain_10) wins.
scenario_names <- strsplit(params$scenarios, ",", fixed = TRUE)[[1]]
scenario_names <- scenario_names[nzchar(scenario_names)]
scenario_names <- scenario_names[order(nchar(scenario_names), decreasing = TRUE)]

match_scenario <- function(text) {
  out <- rep(NA_character_, length(text))
  for (s in scenario_names) {
    hit <- is.na(out) & grepl(paste0("(^|_)", s, "(_|$)"), text)
    out[hit] <- s
  }
  out
}

# Tool-runner benchmarks carry the scenario in the wildcards
# ({scenario}_{param_id}); per-scenario setup rules carry it in the rule name
# (rule_{scenario}), so try both and strip the suffix from those rule names.
df$scenario <- match_scenario(df$wildcards)
from_rule <- is.na(df$scenario)
df$scenario[from_rule] <- match_scenario(df$rule[from_rule])
df$rule[from_rule] <- mapply(
  function(r, s) if (is.na(s)) r else sub(paste0("_", s, "$"), "", r),
  df$rule[from_rule], df$scenario[from_rule], USE.NAMES = FALSE)

df$param_id <- str_extract(df$wildcards, "mc[0-9.]+(?:_[a-z]+[0-9.]+)*")
```

## Inputs

`r bench_dir` (`r length(files)` files).

`max_rss` comes from snakemake's benchmark TSVs. It is the per-process peak resident set size in MiB, sampled by snakemake every 10 s. Multi-process rules report the parent only, so high-fan-out runs (derfinder loading per-sample BigWigs at once) may under-report.

```{r table, results = "asis"}
cat('<details><summary>Click to expand the full per-invocation table</summary>\n\n')
cat(knitr::kable(
  df %>% select(rule, wildcards, wall_s, max_rss_mb) %>% arrange(rule, wildcards),
  digits = 1, format = "html"
))
cat('\n\n</details>\n\n')
```

## Wall time per rule

Each point is one rule invocation. Shape encodes which tool drove it (circle = fastder, triangle = derfinder, square = megadepth_baseline, diamond = grohmm, x = none of these). Panels split by scenario. The hollow black point marks each rule's median across invocations.

```{r time, fig.width = 9, fig.height = 7}
df_pl <- df %>%
  mutate(scenario_lbl = ifelse(is.na(scenario), "n/a", scenario))
medians_time <- df_pl %>%
  group_by(rule, scenario_lbl) %>%
  summarise(wall_s = median(wall_s, na.rm = TRUE), .groups = "drop")
ggplot(df_pl, aes(x = rule, y = wall_s, shape = tool)) +
  geom_jitter(alpha = 0.6, size = 1.6, width = 0.15, height = 0,
              colour = "grey25") +
  geom_point(data = medians_time, shape = 21, size = 2.4,
             colour = "black", fill = "white", stroke = 0.7) +
  coord_flip() +
  scale_y_continuous(trans = "log10") +
  scale_shape_manual(values = c(fastder = 16, derfinder = 17,
                                megadepth_baseline = 15, grohmm = 18),
                     na.value = 4) +
  facet_wrap(~ scenario_lbl, ncol = 2) +
  labs(x = NULL, y = "Wall time (s, log scale)") +
  pub_theme +
  theme(aspect.ratio = NULL)
```

## Max resident memory per rule

Same encoding as above (shape = tool, panel = scenario, hollow = median).

```{r mem, fig.width = 9, fig.height = 7}
medians_mem <- df_pl %>%
  group_by(rule, scenario_lbl) %>%
  summarise(max_rss_mb = median(max_rss_mb, na.rm = TRUE), .groups = "drop")
ggplot(df_pl, aes(x = rule, y = max_rss_mb, shape = tool)) +
  geom_jitter(alpha = 0.6, size = 1.6, width = 0.15, height = 0,
              colour = "grey25") +
  geom_point(data = medians_mem, shape = 21, size = 2.4,
             colour = "black", fill = "white", stroke = 0.7) +
  coord_flip() +
  scale_y_continuous(trans = "log10") +
  scale_shape_manual(values = c(fastder = 16, derfinder = 17,
                                megadepth_baseline = 15, grohmm = 18),
                     na.value = 4) +
  facet_wrap(~ scenario_lbl, ncol = 2) +
  labs(x = NULL, y = "Max resident set size (MiB, log scale)") +
  pub_theme +
  theme(aspect.ratio = NULL)
```

## Wall time vs max memory (tool runners only)

Each invocation of a tool runner (`run_fastder`, `run_derfinder`, `run_megadepth_baseline`, `run_grohmm`) as a single point coloured by tool. The larger solid point is the median across that tool's invocations.

```{r scatter, fig.width = 7, fig.height = 7}
df_runners <- df %>% filter(rule %in% c("run_fastder", "run_derfinder",
                                        "run_megadepth_baseline", "run_grohmm"))
if (nrow(df_runners) > 0) {
  df_centroids <- df_runners %>%
    group_by(tool) %>%
    summarise(wall_s = median(wall_s, na.rm = TRUE),
              max_rss_mb = median(max_rss_mb, na.rm = TRUE),
              .groups = "drop")
  ggplot(df_runners, aes(x = wall_s, y = max_rss_mb, colour = tool)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_point(data = df_centroids, size = 4, stroke = 0.6,
               colour = "black", aes(fill = tool), shape = 21,
               show.legend = FALSE) +
    scale_colour_manual(values = tool_palette, name = NULL) +
    scale_fill_manual(values = tool_palette, guide = "none") +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10") +
    labs(x = "Wall time (s, log scale)", y = "Max RSS (MiB, log scale)") +
    pub_theme +
    theme(legend.position = "right")
} else {
  cat("no tool runner benchmarks yet")
}
```

# Per-tool footprint

The four tools (`fastder`, `derfinder`, `megadepth_baseline`, `grohmm`) each run in a dedicated rule (`run_<tool>`) and feed two shared per-tool stages (`run_gffcompare`, `eval_fuzzy_metrics`). The plots below keep only the rules that carry a `tool` label, so the comparison stays like-for-like.

```{r tool_subset}
df_tool <- df %>% filter(!is.na(tool))
have_tool_data <- nrow(df_tool) > 0
```

## Tool runner: wall time

```{r tool_runner_time, fig.width = 8, fig.height = 4.5}
if (have_tool_data) {
  d_run <- df_tool %>% filter(rule %in% c("run_fastder", "run_derfinder",
                                          "run_megadepth_baseline", "run_grohmm"))
  if (nrow(d_run) > 0) {
    ggplot(d_run, aes(x = tool, y = wall_s, fill = tool)) +
      geom_violin(alpha = 0.5, scale = "width", trim = TRUE, colour = "grey40") +
      geom_jitter(width = 0.15, size = 1.4, alpha = 0.6) +
      stat_summary(fun = median, geom = "point", shape = 21, size = 3,
                   colour = "black", fill = "white", stroke = 0.7) +
      scale_fill_manual(values = tool_palette, guide = "none") +
      labs(x = NULL, y = "Wall time (s)") +
      pub_theme +
      theme(axis.text.x = element_text(angle = 30, hjust = 1))
  } else {
    cat("no tool runner benchmarks yet")
  }
}
```

## Tool runner: max resident memory

This is the most direct way to spot which tool is memory-hungry. Derfinder loads per-sample coverage Rles for every chromosome it processes and converts them to dense numeric vectors before thresholding. Its peak therefore grows roughly linearly in (number of samples) x (chromosome length).

```{r tool_runner_mem, fig.width = 8, fig.height = 4.5}
if (have_tool_data) {
  d_run <- df_tool %>% filter(rule %in% c("run_fastder", "run_derfinder",
                                          "run_megadepth_baseline", "run_grohmm"))
  if (nrow(d_run) > 0) {
    ggplot(d_run, aes(x = tool, y = max_rss_mb, fill = tool)) +
      geom_violin(alpha = 0.5, scale = "width", trim = TRUE, colour = "grey40") +
      geom_jitter(width = 0.15, size = 1.4, alpha = 0.6) +
      stat_summary(fun = median, geom = "point", shape = 21, size = 3,
                   colour = "black", fill = "white", stroke = 0.7) +
      scale_fill_manual(values = tool_palette, guide = "none") +
      labs(x = NULL, y = "Max RSS (MiB)") +
      pub_theme +
      theme(axis.text.x = element_text(angle = 30, hjust = 1))
  }
}
```

## Tool runner: time vs memory

```{r tool_runner_scatter, fig.width = 9, fig.height = 4}
if (have_tool_data) {
  d_run <- df_tool %>% filter(rule %in% c("run_fastder", "run_derfinder",
                                          "run_megadepth_baseline", "run_grohmm"))
  if (nrow(d_run) > 0) {
    ggplot(d_run, aes(x = wall_s, y = max_rss_mb)) +
      geom_point(size = 2.5, alpha = 0.8, colour = "grey20") +
      facet_wrap(~ tool, ncol = 3) +
      labs(x = "Wall time (s)", y = "Max RSS (MiB)") +
      pub_theme
  }
}
```

## Per-tool summary table

Median, max, and total wall time and max-RSS for the per-tool rules.

```{r tool_summary_table, results = "asis"}
if (have_tool_data) {
  d_run <- df_tool %>% filter(rule %in% c("run_fastder", "run_derfinder",
                                          "run_megadepth_baseline", "run_grohmm"))
  if (nrow(d_run) > 0) {
    summary_tbl <- d_run %>%
      group_by(tool) %>%
      summarise(
        runs = n(),
        wall_s_median = median(wall_s, na.rm = TRUE),
        wall_s_max = max(wall_s, na.rm = TRUE),
        wall_s_total = sum(wall_s, na.rm = TRUE),
        max_rss_mb_median = median(max_rss_mb, na.rm = TRUE),
        max_rss_mb_peak = max(max_rss_mb, na.rm = TRUE),
        .groups = "drop"
      )
    print(knitr::kable(summary_tbl, digits = 1, format = "html"))
  }
}
```
