---
title: "fastder-evaluation: simulation depth sweep"
output:
  html_document:
    toc: true
    toc_float: true
params:
  results_root: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5, fig.align = "center")
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
```

This report stitches the per-depth simulation runs into one view. Each run is
a separate config (`config_full_simulation.yaml` for 10M reads,
`config_full_simulation_<N>M.yaml` for the others) with its own results
directory. The figures show how detection scales with sequencing depth.

```{r load}
root <- params$results_root
run_dirs <- list.dirs(root, recursive = FALSE, full.names = TRUE)
run_dirs <- run_dirs[grepl("^config_full_simulation", basename(run_dirs))]

# The base config is the 10M point; the _<N>M variants name their depth.
depth_of <- function(dir_name) {
  m <- str_match(dir_name, "_([0-9]+)M$")[, 2]
  if (is.na(m)) 10L else as.integer(m)
}

read_run <- function(dir_path, file_name) {
  path <- file.path(dir_path, file_name)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE) %>%
    mutate(depth_M = depth_of(basename(dir_path)))
}

summary_all <- bind_rows(lapply(run_dirs, read_run, "summary.csv"))
distances_all <- bind_rows(lapply(run_dirs, read_run, "fuzzy_distances.csv"))
```

```{r check}
if (is.null(summary_all) || nrow(summary_all) == 0) {
  knitr::knit_exit("No per-depth results found under the results root. Run the
  config_full_simulation configs first (see the Makefile).")
}
knitr::kable(
  summary_all %>% distinct(depth_M) %>% arrange(depth_M),
  caption = "Sequencing depths found", col.names = "depth (M reads/sample)")
```

## Sensitivity and precision versus depth

Transcript-level and exon-level gffcompare metrics, averaged across samples
and parameter combinations, line per tool, faceted by scenario.

```{r sens_prec}
levels_long <- bind_rows(
  summary_all %>% transmute(tool, scenario, depth_M, level = "Transcript",
                            sensitivity = transcript_sens,
                            precision = transcript_prec),
  summary_all %>% transmute(tool, scenario, depth_M, level = "Exon",
                            sensitivity = exon_sens, precision = exon_prec)
) %>%
  pivot_longer(c(sensitivity, precision),
               names_to = "metric", values_to = "value") %>%
  group_by(tool, scenario, depth_M, level, metric) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

ggplot(levels_long, aes(x = depth_M, y = value, colour = tool, shape = tool)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.4) +
  scale_x_continuous(trans = "log10", breaks = sort(unique(levels_long$depth_M))) +
  scale_colour_brewer(palette = "Set2", name = NULL) +
  scale_shape(name = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  facet_grid(level + metric ~ scenario) +
  labs(x = "Reads per sample (M, log scale)", y = "Percent") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

## Boundary precision versus depth

Fraction of predicted exon boundaries within 5 bp of a reference boundary,
averaged across samples and parameter combinations.

```{r boundary}
if (!is.null(distances_all) && nrow(distances_all) > 0) {
  b5 <- distances_all %>%
    mutate(distance = as.integer(distance)) %>%
    group_by(tool, scenario, depth_M, sample, param_id) %>%
    summarise(pct = mean(abs(distance) <= 5) * 100, .groups = "drop") %>%
    group_by(tool, scenario, depth_M) %>%
    summarise(pct = mean(pct), .groups = "drop")
  ggplot(b5, aes(x = depth_M, y = pct, colour = tool, shape = tool)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2.4) +
    scale_x_continuous(trans = "log10", breaks = sort(unique(b5$depth_M))) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    scale_shape(name = NULL) +
    coord_cartesian(ylim = c(0, 100)) +
    facet_wrap(~ scenario) +
    labs(x = "Reads per sample (M, log scale)",
         y = "Predicted boundaries within 5 bp (%)") +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom")
} else {
  cat("No fuzzy_distances data found.")
}
```

```{r session}
sessionInfo()
```
