---
title: "fastder-evaluation: simulation depth sweep"
output:
  html_document:
    toc: true
    toc_float: true
params:
  results_root: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5, fig.align = "center",
                      dev = c("png", "pdf"), fig.path = "meta_figs/")
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

# Tool palette. Same hex values as summary.Rmd and benchmarks.Rmd so the
# depth-sweep figures match the rest of the reports.
tool_palette <- c(
  derfinder          = "#66C2A5",
  fastder            = "#FC8D62",
  grohmm             = "#8DA0CB",
  megadepth_baseline = "#E78AC3"
)
```

This report stitches the per-depth simulation runs into one view. Each run is
a separate config (`config_full_simulation.yaml` for 10M reads,
`config_full_simulation_<N>M.yaml` for the others) with its own results
directory. The figures show how detection scales with sequencing depth. A depth
drawn more than once (`config_full_simulation_rep<N>.yaml`) is shown by its
mean, with a bar from the lowest to the highest replicate.

```{r load}
root <- params$results_root
run_dirs <- list.dirs(root, recursive = FALSE, full.names = TRUE)
run_dirs <- run_dirs[grepl("^config_full_simulation", basename(run_dirs))]

# The base config is the 10M point; the _<N>M variants name their depth.
depth_of <- function(dir_name) {
  m <- str_match(dir_name, "_([0-9]+)M$")[, 2]
  if (is.na(m)) 10L else as.integer(m)
}

# The base config is replicate 1; the _rep<N> variants name theirs.
replicate_of <- function(dir_name) {
  m <- str_match(dir_name, "_rep([0-9]+)$")[, 2]
  if (is.na(m)) 1L else as.integer(m)
}

read_run <- function(dir_path, file_name) {
  path <- file.path(dir_path, file_name)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE) %>%
    mutate(depth_M = depth_of(basename(dir_path)),
           replicate = replicate_of(basename(dir_path)))
}

# Mean, lowest and highest replicate of a value already averaged per replicate.
# A replicate with no value is left out.
across_replicates <- function(df, ...) {
  df %>% filter(!is.na(value)) %>% group_by(...) %>%
    summarise(low = min(value), high = max(value), value = mean(value),
              .groups = "drop")
}

# The fastder grid carries two axes that are not accuracy settings: the
# junction-integration ablation (ns1) and the junction read-support filter
# (mjr). Every chunk below averages over the grid, so leaving them in would
# move the reported numbers with nothing in the plot saying so. They have their
# own figures, built by scripts/collect_param_sweeps.py. Identifiers come from
# scripts/param_grid.py, whose tests pin these two patterns.
# min_coverage is kept at the reference threshold of scripts/param_grid.py, the
# value the tools are compared at; groHMM has no threshold and is kept.
REFERENCE_MIN_COVERAGE <- 0.005
default_grid <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  mjr <- suppressWarnings(as.integer(str_extract(df$param_id, "(?<=mjr)[0-9]+")))
  threshold <- suppressWarnings(as.numeric(str_extract(df$param_id, "(?<=mc)[0-9.]+")))
  df[!grepl("(^|_)ns1(_|$)", df$param_id) & (is.na(mjr) | mjr == 0L) &
       (is.na(threshold) | threshold == REFERENCE_MIN_COVERAGE), ]
}

summary_all <- default_grid(bind_rows(lapply(run_dirs, read_run, "summary.csv")))
distances_all <- default_grid(bind_rows(lapply(run_dirs, read_run, "fuzzy_distances.csv")))
```

```{r check}
if (is.null(summary_all) || nrow(summary_all) == 0) {
  knitr::knit_exit("No per-depth results found under the results root. Run the
  config_full_simulation configs first (see the Makefile).")
}
knitr::kable(
  summary_all %>% distinct(depth_M, replicate) %>% count(depth_M, name = "replicates"),
  caption = "Sequencing depths found",
  col.names = c("depth (M reads/sample)", "replicates"))
```

## Sensitivity and precision versus depth

Transcript-level and exon-level gffcompare metrics, averaged across samples
and parameter combinations, line per tool, faceted by scenario.

```{r sens_prec}
levels_long <- bind_rows(
  summary_all %>% transmute(tool, scenario, depth_M, replicate, level = "Transcript",
                            sensitivity = transcript_sens,
                            precision = transcript_prec),
  summary_all %>% transmute(tool, scenario, depth_M, replicate, level = "Exon",
                            sensitivity = exon_sens, precision = exon_prec)
) %>%
  pivot_longer(c(sensitivity, precision),
               names_to = "metric", values_to = "value") %>%
  group_by(tool, scenario, depth_M, replicate, level, metric) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  across_replicates(tool, scenario, depth_M, level, metric)

ggplot(levels_long, aes(x = depth_M, y = value, colour = tool, shape = tool)) +
  geom_linerange(aes(ymin = low, ymax = high), show.legend = FALSE) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.4) +
  scale_x_continuous(trans = "log10", breaks = sort(unique(levels_long$depth_M))) +
  scale_colour_manual(values = tool_palette, name = NULL) +
  scale_shape(name = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  facet_grid(level + metric ~ scenario) +
  labs(x = "Reads per sample (M, log scale)", y = "Percent") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

## Boundary precision versus depth

Fraction of predicted exon boundaries within 5 bp of a reference boundary,
averaged across samples and parameter combinations.

```{r boundary}
if (!is.null(distances_all) && nrow(distances_all) > 0) {
  b5 <- distances_all %>%
    mutate(distance = as.integer(distance)) %>%
    group_by(tool, scenario, depth_M, replicate, sample, param_id) %>%
    summarise(pct = mean(abs(distance) <= 5) * 100, .groups = "drop") %>%
    group_by(tool, scenario, depth_M, replicate) %>%
    summarise(value = mean(pct), .groups = "drop") %>%
    across_replicates(tool, scenario, depth_M) %>%
    rename(pct = value)
  ggplot(b5, aes(x = depth_M, y = pct, colour = tool, shape = tool)) +
    geom_linerange(aes(ymin = low, ymax = high), show.legend = FALSE) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2.4) +
    scale_x_continuous(trans = "log10", breaks = sort(unique(b5$depth_M))) +
    scale_colour_manual(values = tool_palette, name = NULL) +
    scale_shape(name = NULL) +
    coord_cartesian(ylim = c(0, 100)) +
    facet_wrap(~ scenario) +
    labs(x = "Reads per sample (M, log scale)",
         y = "Predicted boundaries within 5 bp (%)") +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom")
} else {
  cat("No fuzzy_distances data found.")
}
```

```{r session}
sessionInfo()
```
