---
title: "recount3 backend: TDP-43 knockdown versus control"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  manifest_csv: ""
  summary_csv: ""
  reference_gtf: ""
  study: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicRanges)
  library(Gviz)
  library(ggplot2)
})
```

This report compares the two sample groups of recount3 study
`r params$study`: a TDP-43 knockdown group and a scramble control group. The
coverage and splice junctions come from recount3, which holds the
Monorail-processed outputs. Each group was processed on its own, so the
called regions of the knockdown group and the control group can be placed
side by side.

This report is specific to the TDP-43 knockdown experiment. The cryptic exon
panel used below (STMN2, HDGFL2, ELAVL3, CELF5, KCNQ2) is hardcoded for it
(Klim et al. 2019, recount3 study SRP166282), with loci picked from the
study's knockdown-specific novel splice junctions. If the recount3 backend is
run on a different study, the cryptic exon panels still render but are not
meaningful for it.

Loss of TDP-43 lets cryptic exons appear in many genes. The panel here
covers STMN2 on chr8, HDGFL2, ELAVL3 and CELF5 on chr19, and KCNQ2 on chr20.
These cryptic exons are not part of the Ensembl annotation, so they cannot be
scored as true positives. The aim is a visual check: a cryptic exon should
show up as a called region, and usually as coverage, in the knockdown group,
and should be weak or absent in the control group. STMN2 is the strongest
case. The others are weaker and are clearest as a called region rather than
as a coverage peak.

This dataset is analysed by two runs that differ only in the `min_coverage`
threshold. The showcase run (`config_klim_2019_tdp43_recount3.yaml`) uses a
clean 1.0 CPM threshold that isolates STMN2, the one cryptic exon well above
the intronic noise floor. The panel run
(`config_klim_2019_tdp43_recount3_panel.yaml`) uses a low 0.02 CPM threshold
so the weaker cryptic exons are emitted too, at the cost of noise. This
report renders for whichever run produced it; see the README section "TDP-43
cryptic exons: showcase and panel" for the rationale.

```{r load-manifest}
manifest <- read.csv(params$manifest_csv, stringsAsFactors = FALSE)
manifest <- manifest[order(manifest$group, manifest$sample), ]
knitr::kable(manifest[, c("group", "sample")],
             caption = "Samples and their groups", row.names = FALSE)

groups <- unique(manifest$group)
group_colors <- setNames(
  c("#d73027", "#4575b4")[seq_along(groups)], groups)
```

```{r regions}
# Cryptic exon panel for the TDP-43 knockdown, hg38. The plot window is the
# gene body plus a small flank. The cryptic exon coordinates (ce_start,
# ce_end) come from the knockdown-specific novel splice junctions in the
# recount3 junction matrix: for HDGFL2 and ELAVL3 the cryptic exon is the gap
# between a flanking junction pair; for STMN2, CELF5 and KCNQ2 a single
# knockdown-specific junction was seen, so the box is anchored on it and on
# the knockdown coverage peak.
regions <- data.frame(
  gene = c("STMN2", "HDGFL2", "ELAVL3", "CELF5", "KCNQ2"),
  chrom = c("chr8", "chr19", "chr19", "chr19", "chr20"),
  start = c(79606000, 4467000, 11446000, 3219000, 63395000),
  end = c(79671000, 4507000, 11486000, 3302000, 63478000),
  ce_start = c(79616821, 4492015, 11463496, 3278200, 63444558),
  ce_end = c(79617200, 4492152, 11463662, 3278400, 63444758),
  stringsAsFactors = FALSE
)
knitr::kable(regions, caption = "Cryptic exon panel: plot window and expected cryptic exon (hg38)")
```

## Coverage view at the cryptic exon loci

For each gene the panel shows, from top to bottom: the position axis, the
Ensembl gene models, one coverage track per sample (knockdown in red, control
in blue), and the called regions of each tool (fastder, derfinder,
megadepth_baseline, grohmm) for each group. The shaded box marks the expected cryptic
exon. A cryptic exon should appear as a coverage signal inside an intron in
the knockdown samples that is missing in the control samples, with a matching
called region in the knockdown group. The cryptic exon is not in the Ensembl
models, so a called region overlapping the box where Ensembl has no exon is
the signal to look for.

```{r tool-regions}
# The Ensembl GTF uses chromosome names without the "chr" prefix; the BigWigs
# and the tool GTFs use "chr". Rename the annotation to match. The cryptic
# exons are not in Ensembl, so a tool region where Ensembl has no exon is the
# cryptic-exon signal.
gene_models <- NULL
if (nzchar(params$reference_gtf) && file.exists(params$reference_gtf)) {
  gene_models <- tryCatch({
    ann <- import(params$reference_gtf)
    seqlevels(ann) <- paste0("chr", sub("^chr", "", seqlevels(ann)))
    ann
  }, error = function(e) NULL)
}

# Manifest column and track color for each tool.
tools <- c(fastder = "fastder_gtf",
           derfinder = "derfinder_gtf",
           megadepth_baseline = "megadepth_gtf",
           grohmm = "grohmm_gtf")
# Tool palette. Same hex values as summary.Rmd, benchmarks.Rmd and
# summary_custom.Rmd so the TDP-43 panels carry the same tool colours as
# the rest of the reports.
tool_colors <- c(
  derfinder          = "#66C2A5",
  fastder            = "#FC8D62",
  grohmm             = "#8DA0CB",
  megadepth_baseline = "#E78AC3"
)
# Short track-title labels; "megadepth_baseline" is too long for the strip.
tool_labels <- c(fastder = "fastder",
                 derfinder = "derfinder",
                 megadepth_baseline = "megadepth",
                 grohmm = "grohmm")

# Keep the exon features rather than the transcript features. fastder
# emits one transcript per stitched ER, spanning from the first exon's start
# to the last exon's end (intron included). Plotting transcripts collapses
# every multi-exon stitched ER into one wide bar covering its introns, which
# hides whether the cryptic exon was called as its own ER. Plotting exons
# instead shows each ER component as a separate bar, so the cryptic exon
# block is visible. The other tools emit single-exon transcripts, so their
# tracks look the same under either filter.
load_regions <- function(gtf) {
  if (length(gtf) == 0 || !nzchar(gtf) || !file.exists(gtf)) return(GRanges())
  gr <- tryCatch(import(gtf), error = function(e) GRanges())
  if (length(gr) > 0 && "type" %in% colnames(mcols(gr))) {
    ex <- gr[!is.na(gr$type) & gr$type == "exon"]
    if (length(ex) > 0) gr <- ex
  }
  gr
}

# regions_by[[tool]][[group]] holds that tool's called regions for the group.
regions_by <- list()
for (tool in names(tools)) {
  regions_by[[tool]] <- list()
  for (group in groups) {
    gtf <- unique(manifest[[tools[[tool]]]][manifest$group == group])
    regions_by[[tool]][[group]] <- load_regions(if (length(gtf)) gtf[1] else "")
  }
}

# Per-sample CPM scaling. fastder's library_size is the area under the
# coverage curve, sum(width * coverage), and CPM is raw_coverage * 1e6 /
# library_size. Recomputing the same factor from the BigWig keeps the Y
# axis of the coverage tracks in the same units as the --min-coverage
# threshold, so the threshold horizontal line is meaningful.
cpm_scale <- function(bigwig) {
  if (!nzchar(bigwig) || !file.exists(bigwig)) return(1.0)
  gr <- tryCatch(rtracklayer::import(bigwig, as = "GRanges"),
                 error = function(e) NULL)
  if (is.null(gr) || length(gr) == 0) return(1.0)
  lib <- sum(as.numeric(width(gr)) * as.numeric(score(gr)))
  if (!is.finite(lib) || lib <= 0) 1.0 else 1e6 / lib
}
manifest$cpm_factor <- vapply(manifest$bigwig, cpm_scale, numeric(1))

# Coverage threshold used by fastder for this run. Pulled from the GTF
# transcript IDs by parsing the parameter id (e.g. mc0.02_ct1.0_ml10_pt5).
# Falls back to 0.05 (the fastder default) when no GTF is available.
extract_min_cov <- function() {
  for (tool in c("fastder")) {
    for (group in groups) {
      gtf <- unique(manifest[[tools[[tool]]]][manifest$group == group])
      if (length(gtf) && nzchar(gtf[1]) && file.exists(gtf[1])) {
        m <- regmatches(gtf[1], regexpr("mc[0-9.]+", gtf[1]))
        if (length(m) && nchar(m)) {
          return(as.numeric(sub("mc", "", m)))
        }
      }
    }
  }
  0.05
}
min_cov_cpm <- extract_min_cov()
```

```{r coverage-plots, results="asis", fig.width=13, fig.height=13}
for (i in seq_len(nrow(regions))) {
  r <- regions[i, ]
  cat("\n\n### ", r$gene, " (", r$chrom, ":", r$start, "-", r$end, ")\n\n",
      sep = "")
  plotted <- tryCatch({
    axis_track <- GenomeAxisTrack()

    # Tracks the highlight box is drawn across: gene models, coverage, regions.
    # body_sizes gives each track an explicit relative height, so the thin
    # region tracks are not starved of space by the coverage tracks.
    body <- list()
    body_sizes <- numeric(0)

    if (!is.null(gene_models)) {
      in_window <- gene_models[seqnames(gene_models) == r$chrom &
                               start(gene_models) <= r$end &
                               end(gene_models) >= r$start &
                               gene_models$type == "exon"]
      if (length(in_window) > 0) {
        body <- c(body, AnnotationTrack(
          in_window, name = "Ensembl", group = in_window$transcript_id,
          stacking = "squish", fill = "grey60", col = NA))
        body_sizes <- c(body_sizes, 1.4)
      }
    }

    # Coverage, one track per sample. Y axis is CPM, computed using the
    # same library size formula as fastder. A dashed horizontal line marks
    # the min_coverage threshold so the reader can see where the called
    # regions cut off.
    for (j in seq_len(nrow(manifest))) {
      m <- manifest[j, ]
      f <- m$cpm_factor
      body <- c(body, DataTrack(
        range = m$bigwig, genome = "hg38", chromosome = r$chrom,
        name = paste0(m$group, "\n", m$sample, "\nCPM"),
        type = c("histogram", "h"),
        transformation = local({
          factor <- f
          function(x) x * factor
        }),
        baseline = min_cov_cpm,
        col.baseline = "#d73027", lty.baseline = 2, lwd.baseline = 0.7,
        col.histogram = group_colors[m$group],
        fill.histogram = group_colors[m$group]))
      body_sizes <- c(body_sizes, 1.6)
    }

    # Called regions, one track per tool and group.
    for (tool in names(tools)) {
      for (group in groups) {
        gr <- regions_by[[tool]][[group]]
        if (length(gr) > 0) {
          gr <- gr[seqnames(gr) == r$chrom &
                   start(gr) <= r$end & end(gr) >= r$start]
        }
        body <- c(body, AnnotationTrack(
          gr, name = paste0(tool_labels[tool], "\n", group),
          chromosome = r$chrom,
          fill = tool_colors[tool], col = NA, stacking = "squish",
          shape = "box"))
        body_sizes <- c(body_sizes, 1.0)
      }
    }

    # Shade the expected cryptic exon across every track below the axis. The
    # box is widened to four times the exon, so its edges fall in flanking
    # sequence and stay visible instead of being hidden under the coverage
    # peak.
    ce_w <- r$ce_end - r$ce_start
    highlight <- HighlightTrack(
      trackList = body, chromosome = r$chrom,
      start = r$ce_start - 1.5 * ce_w, end = r$ce_end + 1.5 * ce_w,
      col = "#d73027", fill = "#fde0dd")

    plotTracks(list(axis_track, highlight),
               from = r$start, to = r$end, chromosome = r$chrom,
               background.title = "white", col.title = "black",
               col.axis = "black", cex.title = 0.8, title.width = 2.2,
               sizes = c(0.6, body_sizes))
    TRUE
  }, error = function(e) {
    cat("Could not draw this region:", conditionMessage(e), "\n")
    FALSE
  })
}
```

## Sample dissimilarity

To check that the knockdown and control samples differ as expected, coverage
is averaged in 50 kb bins over the analysed chromosomes, then compared
between every pair of samples. If the knockdown affects the transcriptome,
the knockdown samples should be more similar to each other than to the
control samples.

```{r dissimilarity}
bin_width <- 50000

dissim <- tryCatch({
  analysed_chroms <- unique(regions$chrom)
  # Bin the analysed chromosomes and read each sample's mean coverage per bin
  # straight from the BigWig. summary() on a BigWigFile does the binning
  # internally, which avoids building genome-wide coverage vectors by hand.
  si <- seqinfo(BigWigFile(manifest$bigwig[1]))
  si <- si[intersect(analysed_chroms, seqnames(si))]
  bins <- unlist(tile(as(si, "GRanges"), width = bin_width))

  cov_mat <- sapply(manifest$bigwig, function(bw) {
    per_bin <- summary(BigWigFile(bw), bins, type = "mean",
                       size = 1L, defaultValue = 0)
    unlist(per_bin)$score
  })
  colnames(cov_mat) <- manifest$sample
  cov_mat <- cov_mat[stats::complete.cases(cov_mat), , drop = FALSE]
  list(cor = cor(cov_mat, method = "spearman"), n_bins = nrow(cov_mat))
}, error = function(e) {
  message("dissimilarity failed: ", conditionMessage(e))
  NULL
})
```

```{r dissimilarity-plot}
if (!is.null(dissim)) {
  cor_mat <- dissim$cor
  long <- expand.grid(s1 = rownames(cor_mat), s2 = colnames(cor_mat))
  long$correlation <- as.vector(cor_mat)
  group_of <- setNames(manifest$group, manifest$sample)
  long$s1 <- factor(long$s1, levels = manifest$sample)
  long$s2 <- factor(long$s2, levels = manifest$sample)

  print(
    ggplot(long, aes(s1, s2, fill = correlation)) +
      geom_tile() +
      geom_text(aes(label = sprintf("%.3f", correlation)), size = 3) +
      scale_fill_gradient(low = "white", high = "#4575b4", limits = c(NA, 1)) +
      coord_fixed() +
      labs(x = NULL, y = NULL,
           title = "Spearman correlation of 50 kb binned coverage",
           subtitle = paste(dissim$n_bins, "bins")) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            aspect.ratio = 1)
  )

  dendro <- hclust(as.dist(1 - cor_mat))
  plot(dendro, main = "Sample clustering (1 - Spearman correlation)",
       xlab = "", sub = "")
} else {
  cat("The dissimilarity summary could not be computed for this run.\n")
}
```

## gffcompare metrics per group

Each group's called regions are compared against the Ensembl annotation with
gffcompare. The TDP-43 cryptic exons are not in that annotation, so they
lower precision rather than raise sensitivity. Read these numbers as a
general agreement check, not as a verdict on the cryptic exons.

```{r summary-table}
if (nzchar(params$summary_csv) && file.exists(params$summary_csv)) {
  summ <- read.csv(params$summary_csv, stringsAsFactors = FALSE)
  cols <- intersect(
    c("tool", "scenario", "param_id", "transcript_sens", "transcript_prec",
      "exon_sens", "exon_prec", "base_sens", "base_prec"),
    colnames(summ))
  knitr::kable(summ[, cols],
               caption = "gffcompare metrics (scenario is the sample group)",
               row.names = FALSE)
} else {
  cat("summary.csv was not available.\n")
}
```

```{r session}
sessionInfo()
```
