---
title: "recount3 backend: TDP-43 knockdown versus control"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  manifest_csv: ""
  summary_csv: ""
  reference_gtf: ""
  study: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicRanges)
  library(Gviz)
  library(ggplot2)
})
```

This report compares the two sample groups of recount3 study
`r params$study`: a TDP-43 knockdown group and a scramble control group. The
coverage and splice junctions come from recount3, which holds the
Monorail-processed outputs. Each group was processed on its own, so the
called regions of the knockdown group and the control group can be placed
side by side.

Loss of TDP-43 is known to let two cryptic exons appear, one in STMN2 on
chr8 and one in UNC13A on chr19. These cryptic exons are not part of the
Ensembl annotation, so they cannot be scored as true positives. The aim here
is a simple visual check: a cryptic exon should show up as coverage and as a
called region in the knockdown group, and should be weak or absent in the
control group.

```{r load-manifest}
manifest <- read.csv(params$manifest_csv, stringsAsFactors = FALSE)
manifest <- manifest[order(manifest$group, manifest$sample), ]
knitr::kable(manifest[, c("group", "sample")],
             caption = "Samples and their groups", row.names = FALSE)

groups <- unique(manifest$group)
group_colors <- setNames(
  c("#d73027", "#4575b4")[seq_along(groups)], groups)
```

```{r regions}
# Cryptic exon loci, hg38. The window has some flank so the gene context is
# visible. Coordinates are approximate gene spans, not exact exon positions.
regions <- data.frame(
  gene  = c("STMN2", "UNC13A"),
  chrom = c("chr8", "chr19"),
  start = c(79590000, 17580000),
  end   = c(79690000, 17710000),
  stringsAsFactors = FALSE
)
knitr::kable(regions, caption = "Plotted regions (hg38)")
```

## Coverage view at the cryptic exon loci

For each locus the panel shows, from top to bottom: the position axis, the
Ensembl gene models, one coverage track per sample (knockdown in red,
control in blue), and the fastder called regions for each group. A cryptic
exon should appear as a coverage signal inside an intron in the knockdown
samples that is missing in the control samples, with a matching fastder
region in the knockdown group.

```{r gene-models}
# The Ensembl GTF uses chromosome names without the "chr" prefix; the BigWigs
# and the fastder GTF use "chr". Rename the annotation to match.
gene_models <- NULL
if (nzchar(params$reference_gtf) && file.exists(params$reference_gtf)) {
  gene_models <- tryCatch({
    ann <- import(params$reference_gtf)
    seqlevels(ann) <- paste0("chr", sub("^chr", "", seqlevels(ann)))
    ann
  }, error = function(e) NULL)
}

fastder_regions <- function(group) {
  gtf <- unique(manifest$fastder_gtf[manifest$group == group])
  gtf <- gtf[nzchar(gtf) & file.exists(gtf)]
  if (length(gtf) == 0) return(GRanges())
  tryCatch(import(gtf[1]), error = function(e) GRanges())
}
fastder_by_group <- lapply(groups, fastder_regions)
names(fastder_by_group) <- groups
```

```{r coverage-plots, results="asis"}
for (i in seq_len(nrow(regions))) {
  r <- regions[i, ]
  cat("\n\n### ", r$gene, " (", r$chrom, ":", r$start, "-", r$end, ")\n\n",
      sep = "")
  plotted <- tryCatch({
    tracks <- list(GenomeAxisTrack())

    if (!is.null(gene_models)) {
      in_window <- gene_models[seqnames(gene_models) == r$chrom &
                               start(gene_models) <= r$end &
                               end(gene_models) >= r$start &
                               gene_models$type == "exon"]
      if (length(in_window) > 0) {
        tracks <- c(tracks, AnnotationTrack(
          in_window, name = "Ensembl", group = in_window$transcript_id,
          stacking = "squish", fill = "grey60", col = NA))
      }
    }

    for (j in seq_len(nrow(manifest))) {
      m <- manifest[j, ]
      tracks <- c(tracks, DataTrack(
        range = m$bigwig, genome = "hg38", chromosome = r$chrom,
        name = paste0(m$sample, "\n", m$group), type = "histogram",
        col.histogram = group_colors[m$group],
        fill.histogram = group_colors[m$group]))
    }

    for (group in groups) {
      gr <- fastder_by_group[[group]]
      if (length(gr) > 0) {
        gr <- gr[seqnames(gr) == r$chrom &
                 start(gr) <= r$end & end(gr) >= r$start]
      }
      tracks <- c(tracks, AnnotationTrack(
        gr, name = paste0("fastder\n", group), chromosome = r$chrom,
        fill = group_colors[group], col = NA, stacking = "dense"))
    }

    plotTracks(tracks, from = r$start, to = r$end, chromosome = r$chrom,
               background.title = "white", col.title = "black",
               col.axis = "black", sizes = NULL)
    TRUE
  }, error = function(e) {
    cat("Could not draw this region:", conditionMessage(e), "\n")
    FALSE
  })
}
```

## Sample dissimilarity

To check that the knockdown and control samples differ as expected, coverage
is averaged in 50 kb bins over the analysed chromosomes, then compared
between every pair of samples. If the knockdown affects the transcriptome,
the knockdown samples should be more similar to each other than to the
control samples.

```{r dissimilarity}
bin_width <- 50000

dissim <- tryCatch({
  analysed_chroms <- unique(regions$chrom)
  # Chromosome lengths come from the first BigWig, so the bins stay within
  # bounds. The whole analysed chromosomes are used, not only the plotted
  # windows.
  si <- seqinfo(BigWigFile(manifest$bigwig[1]))
  si <- si[intersect(analysed_chroms, seqnames(si))]
  region <- as(si, "GRanges")
  bins <- unlist(tile(region, width = bin_width))

  cov_mat <- sapply(seq_len(nrow(manifest)), function(j) {
    gr <- import.bw(manifest$bigwig[j], which = region)
    seqinfo(gr) <- merge(seqinfo(gr), si)[seqnames(seqinfo(gr))]
    cov <- coverage(gr, weight = "score")
    binnedAverage(bins, cov, "mean_cov")$mean_cov
  })
  colnames(cov_mat) <- manifest$sample
  cov_mat <- cov_mat[stats::complete.cases(cov_mat), , drop = FALSE]
  list(cor = cor(cov_mat, method = "spearman"), n_bins = nrow(cov_mat))
}, error = function(e) {
  message("dissimilarity failed: ", conditionMessage(e))
  NULL
})
```

```{r dissimilarity-plot}
if (!is.null(dissim)) {
  cor_mat <- dissim$cor
  long <- expand.grid(s1 = rownames(cor_mat), s2 = colnames(cor_mat))
  long$correlation <- as.vector(cor_mat)
  group_of <- setNames(manifest$group, manifest$sample)
  long$s1 <- factor(long$s1, levels = manifest$sample)
  long$s2 <- factor(long$s2, levels = manifest$sample)

  print(
    ggplot(long, aes(s1, s2, fill = correlation)) +
      geom_tile() +
      geom_text(aes(label = sprintf("%.3f", correlation)), size = 3) +
      scale_fill_gradient(low = "white", high = "#4575b4", limits = c(NA, 1)) +
      labs(x = NULL, y = NULL,
           title = "Spearman correlation of 50 kb binned coverage",
           subtitle = paste(dissim$n_bins, "bins")) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )

  dendro <- hclust(as.dist(1 - cor_mat))
  plot(dendro, main = "Sample clustering (1 - Spearman correlation)",
       xlab = "", sub = "")
} else {
  cat("The dissimilarity summary could not be computed for this run.\n")
}
```

## gffcompare metrics per group

Each group's called regions are compared against the Ensembl annotation with
gffcompare. The TDP-43 cryptic exons are not in that annotation, so they
lower precision rather than raise sensitivity. Read these numbers as a
general agreement check, not as a verdict on the cryptic exons.

```{r summary-table}
if (nzchar(params$summary_csv) && file.exists(params$summary_csv)) {
  summ <- read.csv(params$summary_csv, stringsAsFactors = FALSE)
  cols <- intersect(
    c("tool", "scenario", "param_id", "transcript_sens", "transcript_prec",
      "exon_sens", "exon_prec", "base_sens", "base_prec"),
    colnames(summ))
  knitr::kable(summ[, cols],
               caption = "gffcompare metrics (scenario is the sample group)",
               row.names = FALSE)
} else {
  cat("summary.csv was not available.\n")
}
```

```{r session}
sessionInfo()
```
