---
title: "fastder-evaluation summary"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  summary_csv: NULL
  chain_stats_csv: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 6,
  fig.align = "center"
)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load}
summary_path <- params$summary_csv
chain_path   <- params$chain_stats_csv
if (is.null(summary_path) || !file.exists(summary_path)) {
  stop("summary_csv parameter must point to an existing summary.csv")
}
if (is.null(chain_path) || !file.exists(chain_path)) {
  stop("chain_stats_csv parameter must point to an existing chain_stats.csv")
}
df <- read_csv(summary_path, show_col_types = FALSE)
cs <- read_csv(chain_path, show_col_types = FALSE)
cs$n_exons <- as.integer(cs$n_exons)
cs$total_exon_length <- as.numeric(cs$total_exon_length)
cs$score <- as.numeric(cs$score)
```

# gffcompare metrics

`r summary_path`

```{r summary_table}
knitr::kable(df, digits = 2)
```

## Sensitivity and precision across all levels

Every row in `summary.csv` has `<level>_sens` and `<level>_prec` for the six
gffcompare levels. Plot below shows them all on the same axes per sample.

```{r levels_scatter, fig.width = 6, fig.height = 6}
levels_long <- df %>%
  pivot_longer(
    cols = matches("_(sens|prec)$"),
    names_to = c("level", "metric"),
    names_pattern = "(.+)_(sens|prec)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = metric, values_from = value)

ggplot(levels_long, aes(x = sens, y = prec, colour = level, shape = sample)) +
  geom_point(size = 3) +
  coord_fixed(xlim = c(0, 100), ylim = c(0, 100)) +
  theme_bw()
```

## Per-level bars

```{r level_bars, fig.width = 6, fig.height = 6}
levels_long_metric <- df %>%
  pivot_longer(
    cols = matches("_(sens|prec)$"),
    names_to = c("level", "metric"),
    names_pattern = "(.+)_(sens|prec)",
    values_to = "value"
  )

ggplot(levels_long_metric, aes(x = level, y = value, fill = metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ sample) +
  coord_flip() +
  theme_bw()
```

## Missed and novel features

```{r missed_novel, fig.width = 6, fig.height = 6}
mn <- df %>%
  pivot_longer(
    cols = matches("_(pct|count|total)$"),
    names_to = c("kind", "feature", "field"),
    names_pattern = "(missed|novel)_(exons|introns|loci)_(pct|count|total)",
    values_to = "value"
  ) %>%
  filter(field == "pct")

ggplot(mn, aes(x = feature, y = value, fill = kind)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ sample) +
  coord_flip() +
  theme_bw()
```

## Matching counts

```{r matching_counts}
m_cols <- grep("^matching_", names(df), value = TRUE)
if (length(m_cols) > 0) {
  m <- df %>% select(sample, param_id, all_of(m_cols))
  knitr::kable(m, digits = 0)
}
```

## Query and reference scope

```{r scope_table}
scope_cols <- intersect(
  c("query_mrnas", "query_loci", "query_multi_exon",
    "ref_mrnas", "ref_loci", "ref_multi_exon", "super_loci"),
  names(df)
)
if (length(scope_cols) > 0) {
  knitr::kable(df %>% select(sample, param_id, all_of(scope_cols)), digits = 0)
}
```

# Stitched chain stats

`r chain_path` (`r nrow(cs)` transcripts across `r length(unique(cs$param_id))` parameter combinations)

```{r chain_summary_table}
chain_summary <- cs %>%
  group_by(param_id) %>%
  summarise(
    n_transcripts = n(),
    pct_singletons = mean(n_exons == 1) * 100,
    median_exons = median(n_exons),
    mean_exons = mean(n_exons),
    max_exons = max(n_exons),
    median_length_bp = median(total_exon_length),
    median_score = median(score)
  )
knitr::kable(chain_summary, digits = 2)
```

## Number of exons per transcript

```{r n_exons_hist, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = n_exons)) +
  geom_histogram(binwidth = 1, boundary = 0.5) +
  facet_wrap(~ param_id) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  theme_bw()
```

## Per-chromosome transcript count

```{r per_chrom_count, fig.width = 6, fig.height = 6}
chrom_counts <- cs %>% count(param_id, chrom)
ggplot(chrom_counts, aes(x = chrom, y = n)) +
  geom_col() +
  facet_wrap(~ param_id) +
  coord_flip() +
  theme_bw()
```

## Total exonic length per transcript

```{r length_hist, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = total_exon_length)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ param_id) +
  scale_x_log10() +
  theme_bw()
```

## Score (mean coverage) vs number of exons

```{r score_vs_n_exons, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = n_exons, y = score)) +
  geom_point(alpha = 0.3) +
  scale_y_log10() +
  facet_wrap(~ param_id) +
  theme_bw()
```

## Score vs total exonic length

```{r score_vs_length, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = total_exon_length, y = score)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ param_id) +
  theme_bw()
```
