---
title: "fastder-evaluation run summary (no truth set)"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  summary_csv: NULL
  chain_stats_csv: NULL
  truth_chain_stats_csv: NULL
  fuzzy_jaccard_csv: NULL
  fuzzy_distances_csv: NULL
  fuzzy_recall_csv: NULL
  fuzzy_strand_csv: NULL
  fastder_gtf: NULL
  derfinder_gtf: NULL
  megadepth_gtf: NULL
  truth_gff: NULL
  track_scenario: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicRanges)
})

tool_levels <- c("megadepth_baseline", "derfinder", "fastder")
tool_palette <- c(megadepth_baseline = "#66c2a5",
                  derfinder = "#fc8d62",
                  fastder = "#8da0cb")

# Read a CSV if it exists and is non-empty, otherwise return NULL so each
# section can skip cleanly.
read_opt <- function(path) {
  if (is.null(path) || !file.exists(path)) return(NULL)
  out <- suppressWarnings(readr::read_csv(path, show_col_types = FALSE))
  if (nrow(out) == 0) NULL else out
}

summary_df <- read_opt(params$summary_csv)
chain_df <- read_opt(params$chain_stats_csv)
if (!is.null(chain_df)) {
  chain_df <- chain_df %>%
    mutate(n_exons = as.integer(n_exons),
           total_exon_length = as.numeric(total_exon_length),
           score = as.numeric(score))
}
```

This is the summary report for a run **without a simulated truth set**: real
data from recount3, SRA, or local FASTQ files. There is no known set of true
transcripts to grade precision and recall against, so this report describes
what each tool called and how the tool calls relate to each other, rather than
how correct they are. The biological coverage view for recount3 runs is in the
separate `recount3.html` report.

# Run overview

```{r overview}
if (!is.null(summary_df)) {
  ov <- tibble::tibble(
    field = c("tools", "scenario groups", "samples", "parameter ids"),
    value = c(
      paste(sort(unique(summary_df$tool)), collapse = ", "),
      paste(sort(unique(summary_df$scenario)), collapse = ", "),
      paste(sort(unique(summary_df$sample)), collapse = ", "),
      paste(sort(unique(summary_df$param_id)), collapse = ", ")))
  knitr::kable(ov)
} else {
  cat("no summary.csv data")
}
```

# Called-region counts per tool

`query_loci`, `query_mrnas`, and `query_multi_exon` come from gffcompare's
description of each tool's own output GTF, so they need no truth set.
`query_multi_exon` is the count of transcripts with more than one exon: the
coverage-only baselines (derfinder, megadepth_baseline) emit single-exon
regions, while fastder stitches across splice junctions, so this column is
where the splice-aware caller separates from the baselines.

```{r region_counts}
if (!is.null(summary_df)) {
  cnt <- summary_df %>%
    group_by(tool, scenario, param_id) %>%
    summarise(query_loci = dplyr::first(query_loci),
              query_mrnas = dplyr::first(query_mrnas),
              query_multi_exon = dplyr::first(query_multi_exon),
              .groups = "drop") %>%
    mutate(tool = factor(tool, levels = tool_levels))
  knitr::kable(cnt %>% arrange(scenario, tool, param_id))

  cnt_long <- cnt %>%
    pivot_longer(c(query_loci, query_mrnas, query_multi_exon),
                 names_to = "metric", values_to = "n") %>%
    mutate(n = tidyr::replace_na(n, 0))
  ggplot(cnt_long, aes(tool, n, fill = tool)) +
    geom_col() +
    facet_grid(metric ~ scenario, scales = "free_y") +
    scale_fill_manual(values = tool_palette) +
    labs(x = NULL, y = "count") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
} else {
  cat("no summary.csv data")
}
```

# fastder region structure

`chain_stats.csv` holds one row per region called by fastder, so the next
three sections describe the fastder output directly. The baselines are not
covered here because they do not produce splice chains.

Exons per region: how often fastder stitches more than one exon together.

```{r fastder_exons}
if (!is.null(chain_df)) {
  chain_df %>%
    mutate(exon_bin = factor(pmin(n_exons, 6L),
                             levels = 1:6,
                             labels = c("1", "2", "3", "4", "5", "6+"))) %>%
    ggplot(aes(exon_bin)) +
    geom_bar(fill = tool_palette[["fastder"]]) +
    facet_wrap(~ scenario) +
    labs(x = "exons per region (6 or more collapsed)",
         y = "fastder regions") +
    theme_bw()
} else {
  cat("no chain_stats.csv data")
}
```

Exonic length per region.

```{r fastder_length}
if (!is.null(chain_df)) {
  ggplot(chain_df, aes(total_exon_length)) +
    geom_histogram(bins = 50, fill = tool_palette[["fastder"]]) +
    scale_x_log10() +
    facet_wrap(~ scenario) +
    labs(x = "total exonic length (bp, log10 scale)",
         y = "fastder regions") +
    theme_bw()
} else {
  cat("no chain_stats.csv data")
}
```

Per-scenario summary: region count, multi-exon share, and the median region
size and exon count.

```{r fastder_structure_table}
if (!is.null(chain_df)) {
  chain_df %>%
    group_by(scenario, param_id) %>%
    summarise(n_regions = n(),
              multi_exon = sum(n_exons > 1L),
              multi_exon_pct = round(100 * sum(n_exons > 1L) / n(), 1),
              median_exons = median(n_exons),
              median_length = median(total_exon_length),
              .groups = "drop") %>%
    arrange(scenario, param_id) %>%
    knitr::kable()
} else {
  cat("no chain_stats.csv data")
}
```

# Genomic distribution of fastder regions

Region counts per chromosome, split by strand. Unstranded calls show as `.`;
recount3 hosts unstranded coverage, so recount3 runs are expected to be
unstranded.

```{r fastder_genomic}
if (!is.null(chain_df)) {
  ggplot(chain_df, aes(chrom, fill = strand)) +
    geom_bar() +
    facet_wrap(~ scenario) +
    labs(x = NULL, y = "fastder regions", fill = "strand") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("no chain_stats.csv data")
}
```

# fastder score distribution

The StitchedER score that fastder assigns each region.

```{r fastder_score}
if (!is.null(chain_df)) {
  ggplot(chain_df, aes(score)) +
    geom_histogram(bins = 50, fill = tool_palette[["fastder"]]) +
    facet_wrap(~ scenario, scales = "free_y") +
    labs(x = "fastder StitchedER score", y = "fastder regions") +
    theme_bw()
} else {
  cat("no chain_stats.csv data")
}
```

# Cross-tool agreement

Without a truth set, the comparison between tools is agreement, not accuracy:
how much of the genome each pair of tools calls in common. `jaccard` is the
shared exonic base pairs divided by the union; `frac_a_in_b` is the fraction of
tool A's exonic base pairs that tool B also calls. This is computed on the
representative scenario and first parameter combination per tool (the GTFs
passed to the report), so it is a snapshot, not the full grid.

```{r cross_tool}
import_exons <- function(path) {
  if (is.null(path) || !file.exists(path)) return(NULL)
  gr <- tryCatch(rtracklayer::import(path), error = function(e) NULL)
  if (is.null(gr) || length(gr) == 0) return(NULL)
  ex <- gr[!is.na(gr$type) & gr$type == "exon"]
  if (length(ex) == 0) ex <- gr
  GenomicRanges::reduce(GenomicRanges::granges(ex), ignore.strand = TRUE)
}

gtfs <- list(fastder = import_exons(params$fastder_gtf),
             derfinder = import_exons(params$derfinder_gtf),
             megadepth_baseline = import_exons(params$megadepth_gtf))
gtfs <- gtfs[!vapply(gtfs, is.null, logical(1))]

if (length(gtfs) >= 2) {
  pairs <- utils::combn(names(gtfs), 2, simplify = FALSE)
  ov <- do.call(rbind, lapply(pairs, function(p) {
    a <- gtfs[[p[1]]]
    b <- gtfs[[p[2]]]
    inter <- sum(width(GenomicRanges::intersect(a, b, ignore.strand = TRUE)))
    uni <- sum(width(GenomicRanges::union(a, b, ignore.strand = TRUE)))
    data.frame(tool_a = p[1],
               tool_b = p[2],
               jaccard = round(inter / uni, 3),
               frac_a_in_b = round(inter / sum(width(a)), 3),
               frac_b_in_a = round(inter / sum(width(b)), 3))
  }))
  knitr::kable(ov)
} else {
  cat("fewer than two tool GTFs available, nothing to compare")
}
```
