---
title: "fastder-evaluation summary"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  summary_csv: NULL
  chain_stats_csv: NULL
  fuzzy_jaccard_csv: NULL
  fuzzy_distances_csv: NULL
  fuzzy_recall_csv: NULL
  fuzzy_strand_csv: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 6,
  fig.align = "center"
)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

as_event_levels <- c("es", "mes", "afe", "ale", "mixed")
as_event_labels <- c(
  es = "exon skip (ES)",
  mes = "mutually exclusive (MES)",
  afe = "alt first exon (AFE)",
  ale = "alt last exon (ALE)",
  mixed = "mixed events"
)
as_event_palette <- c(
  es = "#1b9e77",
  mes = "#d95f02",
  afe = "#7570b3",
  ale = "#e7298a",
  mixed = "#666666"
)

label_as <- function(v) {
  factor(v, levels = as_event_levels, labels = as_event_labels[as_event_levels])
}

pub_theme <- theme_bw(base_size = 12) +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey95", colour = NA),
    plot.title = element_text(size = 13, face = "plain")
  )
```

```{r load}
summary_path <- params$summary_csv
chain_path   <- params$chain_stats_csv
if (is.null(summary_path) || !file.exists(summary_path)) {
  stop("summary_csv parameter must point to an existing summary.csv")
}
if (is.null(chain_path) || !file.exists(chain_path)) {
  stop("chain_stats_csv parameter must point to an existing chain_stats.csv")
}
df <- read_csv(summary_path, show_col_types = FALSE)
cs <- read_csv(chain_path, show_col_types = FALSE)
cs$n_exons <- as.integer(cs$n_exons)
cs$total_exon_length <- as.numeric(cs$total_exon_length)
cs$score <- as.numeric(cs$score)

fuzzy_paths <- list(
  jaccard = params$fuzzy_jaccard_csv,
  distances = params$fuzzy_distances_csv,
  recall = params$fuzzy_recall_csv,
  strand = params$fuzzy_strand_csv
)
fuzzy <- lapply(fuzzy_paths, function(p) {
  if (is.null(p) || !file.exists(p)) NULL else read_csv(p, show_col_types = FALSE)
})
```

# Method performance across AS event types

Each ASimulatoR sample carries one alternative-splicing event class: exon skipping (ES), mutually exclusive exons (MES), alternative first exon (AFE), alternative last exon (ALE), or a 4-way mix. The four figures below summarize how fastder reconstructs each class.

## Per-event Jaccard against best reference partner

```{r pub_jaccard, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0) {
  fj <- fuzzy$jaccard
  fj$jaccard <- as.numeric(fj$jaccard)
  fj$as_event <- label_as(fj$sample)
  ggplot(fj, aes(x = as_event, y = jaccard, fill = sample)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.15, outlier.size = 0.5, fill = "white") +
    scale_fill_manual(values = as_event_palette, guide = "none") +
    labs(x = NULL, y = "exonic Jaccard (best fastder match)") +
    coord_cartesian(ylim = c(0, 1)) +
    pub_theme +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("no fuzzy_jaccard data")
}
```

## Per-event boundary precision

Distance from each fastder exon's start to the nearest reference exon start (left), and from end to nearest end (right). Same strand. Distances are clipped to ±200 bp for display.

```{r pub_boundary, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances
  fd$distance <- as.integer(fd$distance)
  fd$as_event <- label_as(fd$sample)
  fd_clip <- fd %>% mutate(distance = pmax(pmin(distance, 200), -200))
  ggplot(fd_clip, aes(x = distance, colour = sample)) +
    geom_density(adjust = 0.6, linewidth = 0.7) +
    scale_colour_manual(values = as_event_palette, name = NULL,
                        labels = as_event_labels[as_event_levels]) +
    facet_wrap(~ boundary) +
    labs(x = "fastder boundary minus nearest ref boundary (bp)", y = "density") +
    pub_theme +
    theme(legend.position = "bottom")
} else {
  cat("no fuzzy_distances data")
}
```

## Per-event strand concordance

```{r pub_strand, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs$category <- factor(fs$category, levels = c("concordant", "discordant", "unstranded", "unmatched"))
  fs$as_event <- label_as(fs$sample)
  fs_pct <- fs %>%
    group_by(sample, as_event, param_id) %>%
    mutate(pct = n_fastder_transcripts / sum(n_fastder_transcripts) * 100) %>%
    ungroup()
  ggplot(fs_pct, aes(y = as_event, x = pct, fill = category)) +
    geom_col() +
    scale_fill_brewer(palette = "Set2", name = NULL) +
    labs(x = "fraction of fastder transcripts (%)", y = NULL) +
    pub_theme +
    theme(aspect.ratio = NULL, legend.position = "bottom")
} else {
  cat("no fuzzy_strand data")
}
```

## Per-event locus-recall curve

Fraction of reference loci with at least f of their exonic length covered by any fastder exon on the same strand.

```{r pub_recall, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$recall) && nrow(fuzzy$recall) > 0) {
  fr <- fuzzy$recall
  fr$threshold <- as.numeric(fr$threshold)
  fr$recall <- as.numeric(fr$recall)
  fr$as_event <- label_as(fr$sample)
  ggplot(fr, aes(x = threshold, y = recall, colour = sample)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_colour_manual(values = as_event_palette, name = NULL,
                        labels = as_event_labels[as_event_levels]) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "exonic coverage threshold f", y = "fraction of ref loci with coverage >= f") +
    pub_theme +
    theme(legend.position = "bottom")
} else {
  cat("no fuzzy_recall data")
}
```

## Per-event scorecard

Headline numbers per AS event class.

```{r pub_scorecard}
if (
  !is.null(fuzzy$jaccard) && !is.null(fuzzy$strand) && !is.null(fuzzy$recall) &&
  nrow(fuzzy$jaccard) > 0 && nrow(fuzzy$strand) > 0 && nrow(fuzzy$recall) > 0
) {
  jrow <- fuzzy$jaccard %>%
    mutate(jaccard = as.numeric(jaccard)) %>%
    group_by(sample) %>%
    summarise(median_jaccard = median(jaccard),
              pct_jaccard_ge_0_5 = mean(jaccard >= 0.5) * 100,
              .groups = "drop")
  srow <- fuzzy$strand %>%
    mutate(n = as.integer(n_fastder_transcripts)) %>%
    pivot_wider(id_cols = sample, names_from = category, values_from = n, values_fill = 0) %>%
    mutate(total = concordant + discordant + unstranded + unmatched,
           pct_concordant = concordant / total * 100,
           pct_discordant = discordant / total * 100) %>%
    select(sample, pct_concordant, pct_discordant)
  rrow <- fuzzy$recall %>%
    mutate(threshold = as.numeric(threshold), recall = as.numeric(recall)) %>%
    filter(threshold == 0.5) %>%
    select(sample, locus_recall_at_0_5 = recall) %>%
    mutate(locus_recall_at_0_5 = locus_recall_at_0_5 * 100)
  scorecard <- jrow %>%
    left_join(srow, by = "sample") %>%
    left_join(rrow, by = "sample") %>%
    mutate(as_event = label_as(sample)) %>%
    select(as_event, median_jaccard, pct_jaccard_ge_0_5, pct_concordant, pct_discordant, locus_recall_at_0_5)
  knitr::kable(scorecard, digits = 2,
               col.names = c("AS event", "median Jaccard",
                             "% Jaccard >= 0.5", "% strand concordant",
                             "% strand discordant", "% loci with >= 50% coverage"))
}
```

# gffcompare metrics

`r summary_path`

```{r summary_table}
knitr::kable(df, digits = 2)
```

## Sensitivity and precision across all levels

Every row in `summary.csv` has `<level>_sens` and `<level>_prec` for the six
gffcompare levels. Plot below shows them all on the same axes per sample.

```{r levels_scatter, fig.width = 6, fig.height = 6}
levels_long <- df %>%
  pivot_longer(
    cols = matches("_(sens|prec)$"),
    names_to = c("level", "metric"),
    names_pattern = "(.+)_(sens|prec)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = metric, values_from = value)

ggplot(levels_long, aes(x = sens, y = prec, colour = level, shape = sample)) +
  geom_point(size = 3) +
  coord_fixed(xlim = c(0, 100), ylim = c(0, 100)) +
  theme_bw()
```

## Per-level bars

```{r level_bars, fig.width = 6, fig.height = 6}
levels_long_metric <- df %>%
  pivot_longer(
    cols = matches("_(sens|prec)$"),
    names_to = c("level", "metric"),
    names_pattern = "(.+)_(sens|prec)",
    values_to = "value"
  )

ggplot(levels_long_metric, aes(x = level, y = value, fill = metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ sample) +
  coord_flip() +
  theme_bw()
```

## Missed and novel features

```{r missed_novel, fig.width = 6, fig.height = 6}
mn <- df %>%
  pivot_longer(
    cols = matches("_(pct|count|total)$"),
    names_to = c("kind", "feature", "field"),
    names_pattern = "(missed|novel)_(exons|introns|loci)_(pct|count|total)",
    values_to = "value"
  ) %>%
  filter(field == "pct")

ggplot(mn, aes(x = feature, y = value, fill = kind)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ sample) +
  coord_flip() +
  theme_bw()
```

## Matching counts

```{r matching_counts}
m_cols <- grep("^matching_", names(df), value = TRUE)
if (length(m_cols) > 0) {
  m <- df %>% select(sample, param_id, all_of(m_cols))
  knitr::kable(m, digits = 0)
}
```

## Query and reference scope

```{r scope_table}
scope_cols <- intersect(
  c("query_mrnas", "query_loci", "query_multi_exon",
    "ref_mrnas", "ref_loci", "ref_multi_exon", "super_loci"),
  names(df)
)
if (length(scope_cols) > 0) {
  knitr::kable(df %>% select(sample, param_id, all_of(scope_cols)), digits = 0)
}
```

# Stitched chain stats

`r chain_path` (`r nrow(cs)` transcripts across `r length(unique(cs$param_id))` parameter combinations)

```{r chain_summary_table}
chain_summary <- cs %>%
  group_by(param_id) %>%
  summarise(
    n_transcripts = n(),
    pct_singletons = mean(n_exons == 1) * 100,
    median_exons = median(n_exons),
    mean_exons = mean(n_exons),
    max_exons = max(n_exons),
    median_length_bp = median(total_exon_length),
    median_score = median(score)
  )
knitr::kable(chain_summary, digits = 2)
```

## Number of exons per transcript

```{r n_exons_hist, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = n_exons)) +
  geom_histogram(binwidth = 1, boundary = 0.5) +
  facet_wrap(~ param_id) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  theme_bw()
```

## Per-chromosome transcript count

```{r per_chrom_count, fig.width = 6, fig.height = 6}
chrom_counts <- cs %>% count(param_id, chrom)
ggplot(chrom_counts, aes(x = chrom, y = n)) +
  geom_col() +
  facet_wrap(~ param_id) +
  coord_flip() +
  theme_bw()
```

## Total exonic length per transcript

```{r length_hist, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = total_exon_length)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ param_id) +
  scale_x_log10() +
  theme_bw()
```

## Score (mean coverage) vs number of exons

```{r score_vs_n_exons, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = n_exons, y = score)) +
  geom_point(alpha = 0.3) +
  scale_y_log10() +
  facet_wrap(~ param_id) +
  theme_bw()
```

## Score vs total exonic length

```{r score_vs_length, fig.width = 6, fig.height = 6}
ggplot(cs, aes(x = total_exon_length, y = score)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ param_id) +
  theme_bw()
```

# Fuzzy fastder vs reference

These metrics drop gffcompare's exact-boundary requirement and look at how much fastder transcripts overlap reference transcripts on the same strand. Useful for fastder because its exon boundaries come from coverage drop-offs, not splice-site coordinates, so it never wins on transcript-exact matching.

## Reciprocal-best Jaccard per reference transcript

For each reference transcript, the highest exonic Jaccard against any fastder transcript on the same strand. A value of 1 means perfect exonic match; 0 means no overlap.

```{r jaccard_hist, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0) {
  fj <- fuzzy$jaccard
  fj$jaccard <- as.numeric(fj$jaccard)
  ggplot(fj, aes(x = jaccard)) +
    geom_histogram(bins = 30, boundary = 0) +
    facet_grid(sample ~ param_id) +
    theme_bw() +
    theme(aspect.ratio = 1, legend.position = "none")
} else {
  cat("no fuzzy_jaccard data")
}
```

```{r jaccard_summary}
if (!is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0) {
  fj <- fuzzy$jaccard
  fj$jaccard <- as.numeric(fj$jaccard)
  jsum <- fj %>%
    group_by(sample, param_id) %>%
    summarise(
      n_ref_transcripts = n(),
      pct_jaccard_zero = mean(jaccard == 0) * 100,
      median_jaccard = median(jaccard),
      mean_jaccard = mean(jaccard),
      pct_jaccard_ge_0_5 = mean(jaccard >= 0.5) * 100,
      pct_jaccard_ge_0_9 = mean(jaccard >= 0.9) * 100,
      .groups = "drop"
    )
  knitr::kable(jsum, digits = 2)
}
```

## Boundary-distance histogram

Signed bp distance from each fastder exon's start to the nearest ref exon start (same strand), and the same for ends. Tight peak at zero means boundaries align; wide spread means they don't.

```{r boundary_hist, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances
  fd$distance <- as.integer(fd$distance)
  fd_clip <- fd %>% mutate(distance = pmax(pmin(distance, 200), -200))
  ggplot(fd_clip, aes(x = distance)) +
    geom_histogram(bins = 80, boundary = 0) +
    facet_grid(sample + param_id ~ boundary) +
    theme_bw() +
    theme(aspect.ratio = 1, legend.position = "none")
} else {
  cat("no fuzzy_distances data")
}
```

```{r boundary_summary}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances
  fd$distance <- as.integer(fd$distance)
  bsum <- fd %>%
    group_by(sample, param_id, boundary) %>%
    summarise(
      n = n(),
      pct_within_5bp = mean(abs(distance) <= 5) * 100,
      pct_within_50bp = mean(abs(distance) <= 50) * 100,
      median_abs = median(abs(distance)),
      .groups = "drop"
    )
  knitr::kable(bsum, digits = 2)
}
```

## Locus recall at fixed coverage thresholds

Fraction of reference loci with at least f of their exonic length covered by any fastder exon on the same strand, for f in {0.1, 0.5, 0.7, 0.9, 0.95}.

```{r locus_recall_curve, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$recall) && nrow(fuzzy$recall) > 0) {
  fr <- fuzzy$recall
  fr$threshold <- as.numeric(fr$threshold)
  fr$recall <- as.numeric(fr$recall)
  ggplot(fr, aes(x = threshold, y = recall)) +
    geom_line() +
    geom_point(size = 2) +
    coord_cartesian(ylim = c(0, 1)) +
    facet_grid(sample ~ param_id) +
    theme_bw() +
    theme(aspect.ratio = 1, legend.position = "none")
} else {
  cat("no fuzzy_recall data")
}
```

```{r locus_recall_table}
if (!is.null(fuzzy$recall) && nrow(fuzzy$recall) > 0) {
  knitr::kable(fuzzy$recall, digits = 3)
}
```

## Strand concordance per fastder transcript

For each fastder StitchedER, the best-overlapping reference transcript (any strand) is selected. The fastder strand is then classified as concordant, discordant, unstranded (fastder emitted '.'), or unmatched (no ref overlap on any strand). Counts and fractions are per (sample, param_id).

```{r strand_counts, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs$category <- factor(fs$category, levels = c("concordant", "discordant", "unstranded", "unmatched"))
  ggplot(fs, aes(x = sample, y = n_fastder_transcripts, fill = category)) +
    geom_col() +
    facet_wrap(~ param_id) +
    theme_bw() +
    theme(aspect.ratio = 1)
} else {
  cat("no fuzzy_strand data")
}
```

```{r strand_pct, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs$category <- factor(fs$category, levels = c("concordant", "discordant", "unstranded", "unmatched"))
  fs_pct <- fs %>%
    group_by(sample, param_id) %>%
    mutate(pct = n_fastder_transcripts / sum(n_fastder_transcripts) * 100) %>%
    ungroup()
  ggplot(fs_pct, aes(x = sample, y = pct, fill = category)) +
    geom_col() +
    facet_wrap(~ param_id) +
    theme_bw() +
    theme(aspect.ratio = 1)
}
```

```{r strand_table}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs_wide <- fs %>%
    pivot_wider(names_from = category, values_from = n_fastder_transcripts, values_fill = 0) %>%
    mutate(total = concordant + discordant + unstranded + unmatched,
           pct_concordant = concordant / total * 100,
           pct_discordant = discordant / total * 100,
           pct_unstranded = unstranded / total * 100,
           pct_unmatched = unmatched / total * 100)
  knitr::kable(fs_wide, digits = 2)
}
```
