---
title: "fastder-evaluation summary"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  summary_csv: NULL
  chain_stats_csv: NULL
  truth_chain_stats_csv: NULL
  fuzzy_jaccard_csv: NULL
  fuzzy_distances_csv: NULL
  fuzzy_recall_csv: NULL
  fuzzy_strand_csv: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 6,
  fig.align = "center"
)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(ggrepel)
})

as_event_levels <- c("es", "mes", "afe", "ale", "mixed")
as_event_labels <- c(
  es = "exon skip (ES)",
  mes = "mutually exclusive (MES)",
  afe = "alt first exon (AFE)",
  ale = "alt last exon (ALE)",
  mixed = "mixed events"
)
as_event_palette <- c(
  es = "#1b9e77",
  mes = "#d95f02",
  afe = "#7570b3",
  ale = "#e7298a",
  mixed = "#666666"
)

label_as <- function(v) {
  factor(v, levels = as_event_levels, labels = as_event_labels[as_event_levels])
}

pub_theme <- theme_bw(base_size = 12) +
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_line(colour = "grey88"),
    panel.grid.minor = element_line(colour = "grey94"),
    strip.background = element_rect(fill = "grey95", colour = NA),
    plot.title = element_text(size = 13, face = "plain")
  )

scenario_labels <- c(
  template_and_variant = "template + variant",
  variant_only = "variant only"
)
relabel_scenario <- function(v) {
  factor(v, levels = names(scenario_labels), labels = scenario_labels[names(scenario_labels)])
}

# Collapsible kable: wraps the rendered table in <details>/<summary> so it
# stays hidden in the report until the reader clicks the summary.
collapsible_kable <- function(df, summary_text = "Click to expand", ...) {
  cat('<details><summary>', summary_text, '</summary>\n\n', sep = "")
  cat(knitr::kable(df, format = "html", ...))
  cat('\n\n</details>\n\n')
}

# Parse the abbreviated param_id (e.g. "mc0.01_ml10_ct0.5") into numeric
# columns for each axis of the parameter grid. Missing axes stay NA.
parse_param_id <- function(df_in) {
  df_in %>%
    mutate(
      min_coverage = suppressWarnings(as.numeric(str_extract(param_id, "(?<=mc)[0-9.]+"))),
      min_length = suppressWarnings(as.numeric(str_extract(param_id, "(?<=ml)[0-9.]+"))),
      coverage_tolerance = suppressWarnings(as.numeric(str_extract(param_id, "(?<=ct)[0-9.]+"))),
      position_tolerance = suppressWarnings(as.numeric(str_extract(param_id, "(?<=pt)[0-9]+")))
    )
}
```

```{r load}
summary_path <- params$summary_csv
chain_path   <- params$chain_stats_csv
if (is.null(summary_path) || !file.exists(summary_path)) {
  stop("summary_csv parameter must point to an existing summary.csv")
}
if (is.null(chain_path) || !file.exists(chain_path)) {
  stop("chain_stats_csv parameter must point to an existing chain_stats.csv")
}
df_all <- read_csv(summary_path, show_col_types = FALSE)
cs <- read_csv(chain_path, show_col_types = FALSE)
cs$n_exons <- as.integer(cs$n_exons)
cs$total_exon_length <- as.numeric(cs$total_exon_length)
cs$score <- as.numeric(cs$score)

truth_chain <- NULL
if (!is.null(params$truth_chain_stats_csv) && file.exists(params$truth_chain_stats_csv)) {
  truth_chain <- read_csv(params$truth_chain_stats_csv, show_col_types = FALSE)
  truth_chain$n_exons <- as.integer(truth_chain$n_exons)
  truth_chain$total_exon_length <- as.numeric(truth_chain$total_exon_length)
}

fuzzy_paths <- list(
  jaccard = params$fuzzy_jaccard_csv,
  distances = params$fuzzy_distances_csv,
  recall = params$fuzzy_recall_csv,
  strand = params$fuzzy_strand_csv
)
fuzzy_all <- lapply(fuzzy_paths, function(p) {
  if (is.null(p) || !file.exists(p)) NULL else read_csv(p, show_col_types = FALSE)
})

# Sections below the "Tool comparison" block are fastder-only: they sweep
# fastder's full parameter grid and facet by AS event class. The
# baselines participate only along the shared axes (min_coverage and, for
# derfinder, position_tolerance) so leaving them in those panels would
# either crowd the figures with sparse points or collapse different
# tools' rows together. Stash a fastder-filtered view here for the rest
# of the report.
df <- df_all %>% filter(tool == "fastder" | is.na(tool))
fuzzy <- lapply(fuzzy_all, function(d) {
  if (is.null(d)) NULL else d %>% filter(tool == "fastder" | is.na(tool))
})
```

# Tool comparison

fastder vs the coverage baselines on the same simulated truth. Shared axes: `mc` (all three tools), `pt` (fastder + derfinder). Each panel below picks the best fastder param combo and matches the baselines at the same `mc` (and `pt` for derfinder).

Note on metric choice. Jaccard, locus recall and base sensitivity reward tools that emit a few large transcripts spanning entire loci, which is what the coverage baselines do by construction. fastder splits each locus into many isoform-like calls, so its best-1 Jaccard against a multi-exon reference is small. The metrics that reflect what fastder actually does live under "Splice-junction-aware metrics" below: boundary precision at small bp tolerances and the count of multi-exon transcripts produced.

```{r tool_setup}
have_tool_data <- !is.null(fuzzy_all$jaccard) && nrow(fuzzy_all$jaccard) > 0 &&
                  !is.null(fuzzy_all$strand) && nrow(fuzzy_all$strand) > 0 &&
                  !is.null(fuzzy_all$recall) && nrow(fuzzy_all$recall) > 0 &&
                  "tool" %in% names(fuzzy_all$jaccard)

extract_mc <- function(pid) suppressWarnings(as.numeric(str_extract(pid, "(?<=mc)[0-9.]+")))
extract_pt <- function(pid) {
  v <- suppressWarnings(as.numeric(str_extract(pid, "(?<=pt)[0-9.]+")))
  ifelse(is.na(v), 0, v)
}

if (have_tool_data) {
  best_fastder_pid <- fuzzy_all$jaccard %>%
    filter(tool == "fastder") %>%
    mutate(jaccard = as.numeric(jaccard)) %>%
    group_by(param_id) %>%
    summarise(median_jaccard = median(jaccard), .groups = "drop") %>%
    slice_max(median_jaccard, n = 1, with_ties = FALSE) %>%
    pull(param_id)
  best_mc <- extract_mc(best_fastder_pid)
  best_pt <- extract_pt(best_fastder_pid)
  best_megadepth_pid <- sprintf("mc%s", format(best_mc, trim = TRUE))
  best_derfinder_pid <- sprintf("mc%s_pt%s", format(best_mc, trim = TRUE),
                                format(best_pt, trim = TRUE))
  pick_tool_param <- function(d) {
    d %>% filter((tool == "fastder" & param_id == best_fastder_pid) |
                 (tool == "megadepth_baseline" & param_id == best_megadepth_pid) |
                 (tool == "derfinder" & param_id == best_derfinder_pid))
  }
}
```

```{r tool_picked, results = "asis"}
if (have_tool_data) {
  cat(sprintf(
    "Selected: fastder `%s`, derfinder `%s`, megadepth_baseline `%s` (shared mc = %s, pt = %s).\n\n",
    best_fastder_pid, best_derfinder_pid, best_megadepth_pid,
    format(best_mc), format(best_pt)
  ))
}
```

## Per-tool best Jaccard

```{r tool_jaccard, fig.width = 10, fig.height = 5}
if (have_tool_data) {
  fj <- fuzzy_all$jaccard %>%
    pick_tool_param() %>%
    mutate(jaccard = as.numeric(jaccard),
           scenario = relabel_scenario(scenario))
  ggplot(fj, aes(x = tool, y = jaccard, fill = tool)) +
    geom_violin(alpha = 0.5, scale = "width", trim = TRUE, colour = "grey40") +
    geom_jitter(width = 0.18, size = 0.25, alpha = 0.2, colour = "grey15") +
    stat_summary(fun = median, geom = "point", shape = 21, size = 2.6,
                 colour = "black", fill = "white", stroke = 0.7) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    facet_wrap(~ scenario, ncol = 2) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = NULL, y = "Best exonic Jaccard against reference") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("no per-tool data: summary.csv lacks a tool column or fuzzy CSVs are empty")
}
```

## Per-tool locus recall at threshold 0.5

Fraction of reference loci with at least 50% of their exonic length covered by any same-strand exon emitted by the tool.

```{r tool_recall, fig.width = 10, fig.height = 5}
if (have_tool_data) {
  fr <- fuzzy_all$recall %>%
    pick_tool_param() %>%
    mutate(threshold = as.numeric(threshold),
           recall = as.numeric(recall),
           scenario = relabel_scenario(scenario)) %>%
    filter(threshold == 0.5)
  ggplot(fr, aes(x = tool, y = recall, fill = tool)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7,
             alpha = 0.85, colour = "grey20") +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    facet_grid(scenario ~ sample) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = NULL, y = "Locus recall at threshold 0.5") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL)
} else {
  cat("no per-tool recall data")
}
```

## Per-tool strand concordance

Fraction of each tool's transcripts whose strand matches the best-overlapping reference transcript on any strand.

```{r tool_strand, fig.width = 10, fig.height = 5}
if (have_tool_data) {
  fs <- fuzzy_all$strand %>%
    pick_tool_param() %>%
    mutate(n = as.integer(n_fastder_transcripts),
           scenario = relabel_scenario(scenario))
  fs_pct <- fs %>%
    group_by(scenario, sample, tool) %>%
    mutate(pct = n / sum(n) * 100) %>%
    ungroup()
  fs_pct$category <- factor(fs_pct$category,
                            levels = c("concordant", "discordant", "unstranded", "unmatched"))
  ggplot(fs_pct, aes(y = tool, x = pct, fill = category)) +
    geom_col() +
    scale_fill_brewer(palette = "Set2", name = NULL) +
    facet_grid(scenario ~ sample) +
    labs(x = "Fraction of tool transcripts (%)", y = NULL) +
    pub_theme +
    theme(aspect.ratio = NULL, legend.position = "bottom")
} else {
  cat("no per-tool strand data")
}
```

## Splice-junction-aware metrics

The metrics above reward tools that emit one large transcript per locus. Below are metrics that need real splice-site structure.

### Boundary precision

Fraction of predicted exon boundaries within each distance threshold. Higher is better. Lines on a single panel per scenario so the gap between fastder and the baselines is read in one glance.

```{r tool_boundary_precision, fig.width = 10, fig.height = 5}
if (have_tool_data && !is.null(fuzzy_all$distances) && nrow(fuzzy_all$distances) > 0) {
  fd <- fuzzy_all$distances %>%
    pick_tool_param() %>%
    mutate(distance = as.integer(distance),
           abs_distance = abs(distance))
  thresholds_bp <- c(0, 1, 5, 20, 50, 200)
  rows <- lapply(thresholds_bp, function(thr) {
    fd %>%
      group_by(tool, scenario) %>%
      summarise(pct = mean(abs_distance <= thr) * 100,
                n = n(), .groups = "drop") %>%
      mutate(threshold_bp = thr)
  })
  fd_pct <- bind_rows(rows) %>%
    mutate(scenario = relabel_scenario(scenario),
           threshold_label = factor(paste0("<=", threshold_bp, " bp"),
                                    levels = paste0("<=", thresholds_bp, " bp")))
  ggplot(fd_pct, aes(x = threshold_bp, y = pct, colour = tool, shape = tool,
                     group = tool)) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    geom_point(size = 3, alpha = 0.9) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    scale_shape_manual(values = c(fastder = 16, derfinder = 17,
                                  megadepth_baseline = 15), name = NULL) +
    scale_x_continuous(breaks = thresholds_bp, trans = "log1p",
                       labels = ifelse(thresholds_bp == 0,
                                       "0 bp",
                                       paste0("<=", thresholds_bp, " bp"))) +
    facet_wrap(~ scenario, ncol = 2) +
    coord_cartesian(ylim = c(0, 100)) +
    labs(x = "Distance threshold", y = "Predicted boundaries within threshold (%)") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL,
          legend.position = "bottom")
} else {
  cat("no per-tool boundary data")
}
```

### Boundary distance ECDF

Cumulative fraction of predicted boundaries within each distance. A steep curve on the left means the tool is hitting splice sites. At `position_tolerance = 0` derfinder reduces to megadepth_baseline so their curves coincide.

```{r tool_boundary_ecdf, fig.width = 9, fig.height = 5}
if (have_tool_data && !is.null(fuzzy_all$distances) && nrow(fuzzy_all$distances) > 0) {
  # Plot order: baseline first, derfinder second, fastder last so all three
  # curves remain visible when two coincide (e.g. derfinder at pt=0 ==
  # megadepth_baseline). The geom_step is drawn in factor order, so the
  # later-listed tool sits on top.
  fd <- fuzzy_all$distances %>%
    pick_tool_param() %>%
    mutate(distance = as.integer(distance),
           abs_distance = pmax(abs(distance), 1L),
           scenario = relabel_scenario(scenario),
           tool = factor(tool, levels = c("megadepth_baseline",
                                          "derfinder", "fastder")))
  ggplot(fd, aes(x = abs_distance, colour = tool, linetype = tool)) +
    stat_ecdf(geom = "step", linewidth = 0.8, alpha = 0.8) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    scale_linetype_manual(values = c(megadepth_baseline = "11",
                                     derfinder = "32",
                                     fastder = "solid"),
                          name = NULL) +
    scale_x_log10(breaks = c(1, 5, 20, 100, 1000, 10000),
                  labels = c("1", "5", "20", "100", "1k", "10k")) +
    coord_cartesian(ylim = c(0, 1)) +
    facet_wrap(~ scenario, ncol = 2) +
    labs(x = "Absolute distance to nearest reference boundary (bp, log scale)",
         y = "Cumulative fraction of predicted boundaries") +
    pub_theme +
    theme(legend.position = "bottom",
          aspect.ratio = NULL)
}
```

### Predicted transcript granularity

Each tool is one point per scenario. x = number of transcripts, y = median transcript length. Top-left = many short calls (fastder, isoform-like); bottom-right = few long calls (coverage baselines).

```{r tool_transcript_granularity, fig.width = 7, fig.height = 5}
if (have_tool_data && !is.null(fuzzy_all$jaccard) && nrow(fuzzy_all$jaccard) > 0) {
  fj <- fuzzy_all$jaccard %>%
    pick_tool_param() %>%
    mutate(fastder_bp = as.numeric(fastder_bp),
           scenario = relabel_scenario(scenario))
  granularity <- fj %>%
    filter(!is.na(fastder_transcript), fastder_transcript != "") %>%
    distinct(tool, scenario, fastder_transcript, fastder_bp) %>%
    group_by(tool, scenario) %>%
    summarise(n_transcripts = n(),
              median_bp = median(fastder_bp, na.rm = TRUE),
              .groups = "drop")
  ggplot(granularity, aes(x = n_transcripts, y = median_bp,
                          colour = tool, shape = tool)) +
    geom_point(size = 4, alpha = 0.85) +
    ggrepel::geom_text_repel(aes(label = tool), size = 3,
                             colour = "grey20",
                             box.padding = 0.5, max.overlaps = Inf) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    scale_shape_manual(values = c(fastder = 16, derfinder = 17,
                                  megadepth_baseline = 15), name = NULL) +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ scenario, ncol = 2) +
    labs(x = "Number of predicted transcripts (log)",
         y = "Median predicted transcript length, bp (log)") +
    pub_theme +
    theme(legend.position = "none")
}
```

### Multi-exon transcripts produced

Each tool is one point per scenario. x = total transcripts, y = multi-exon count. Coverage baselines sit at the bottom (multi-exon = 0 by construction, drawn as 1 on the log scale); fastder is well above.

```{r tool_multiexon, fig.width = 7, fig.height = 5}
if ("tool" %in% names(df_all) && nrow(df_all) > 0) {
  d <- df_all %>%
    filter((tool == "fastder" & param_id == best_fastder_pid) |
           (tool == "megadepth_baseline" & param_id == best_megadepth_pid) |
           (tool == "derfinder" & param_id == best_derfinder_pid)) %>%
    distinct(tool, scenario, param_id, query_mrnas, query_multi_exon) %>%
    mutate(query_mrnas = as.integer(query_mrnas),
           query_multi_exon = as.integer(query_multi_exon),
           scenario = relabel_scenario(scenario)) %>%
    group_by(tool, scenario) %>%
    summarise(query_mrnas = sum(query_mrnas),
              query_multi_exon = sum(query_multi_exon),
              .groups = "drop") %>%
    # log10 cannot show 0; pin baselines to 1 so they sit at the floor of the
    # axis instead of being dropped.
    mutate(query_multi_exon_plot = pmax(query_multi_exon, 1L))
  ggplot(d, aes(x = query_mrnas, y = query_multi_exon_plot,
                colour = tool, shape = tool)) +
    geom_point(size = 4, alpha = 0.85) +
    ggrepel::geom_text_repel(aes(label = tool), size = 3,
                             colour = "grey20",
                             box.padding = 0.5, max.overlaps = Inf) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    scale_shape_manual(values = c(fastder = 16, derfinder = 17,
                                  megadepth_baseline = 15), name = NULL) +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ scenario, ncol = 2) +
    labs(x = "Total predicted transcripts (log)",
         y = "Multi-exon predicted transcripts (log; 0 plotted as 1)") +
    pub_theme +
    theme(legend.position = "none")
}
```

## Tolerance equivalence: fastder pt vs derfinder maxregiongap

Median Jaccard along the position_tolerance / maxregiongap axis at the best mc. fastder slack sits at SJ boundaries; derfinder slack bridges below-threshold gaps. Comparable when the curves sit close.

```{r tool_pt_axis, fig.width = 8, fig.height = 4}
if (have_tool_data) {
  pt_d <- fuzzy_all$jaccard %>%
    filter(tool %in% c("fastder", "derfinder")) %>%
    mutate(jaccard = as.numeric(jaccard),
           mc = extract_mc(param_id),
           pt = extract_pt(param_id)) %>%
    filter(!is.na(mc), abs(mc - best_mc) < 1e-9)
  if (nrow(pt_d) > 0 && length(unique(pt_d$pt)) > 1) {
    pt_summary <- pt_d %>%
      group_by(tool, scenario, pt) %>%
      summarise(median_jaccard = median(jaccard), .groups = "drop") %>%
      mutate(scenario = relabel_scenario(scenario))
    ggplot(pt_summary, aes(x = pt, y = median_jaccard,
                           colour = tool, group = tool)) +
      geom_line(linewidth = 0.7) +
      geom_point(size = 2) +
      scale_colour_brewer(palette = "Set2") +
      facet_wrap(~ scenario, ncol = 2) +
      coord_cartesian(ylim = c(0, 1)) +
      labs(x = "position_tolerance / maxregiongap (bp)",
           y = "Median Jaccard",
           colour = NULL) +
      pub_theme +
      theme(legend.position = "bottom")
  } else {
    cat("only one pt value in the grid; no equivalence sweep to plot")
  }
}
```

## Per-tool gffcompare summary

Sensitivity and precision at every gffcompare level, per tool, at the parameters chosen above.

```{r tool_gffcompare_table, results = "asis"}
if (have_tool_data && nrow(df_all) > 0 && "tool" %in% names(df_all)) {
  d <- df_all %>%
    filter((tool == "fastder" & param_id == best_fastder_pid) |
           (tool == "megadepth_baseline" & param_id == best_megadepth_pid) |
           (tool == "derfinder" & param_id == best_derfinder_pid)) %>%
    select(tool, scenario, sample, param_id,
           matches("_(sens|prec)$"), matches("matching"))
  collapsible_kable(d, "Show per-tool gffcompare table", digits = 2)
}
```

The sections below sweep fastder's parameter grid and facet by AS event class. They are intentionally fastder-only: a parameter that is meaningless for derfinder cannot be plotted on its axis, and the AS event mapping is sample-specific in the same way for every tool.



# Method performance across AS event types

Each ASimulatoR sample carries one alternative-splicing event class: exon skipping (ES), mutually exclusive exons (MES), alternative first exon (AFE), alternative last exon (ALE), or a four-way mix.

Each event class is run through two scenarios that differ in which transcripts contribute reads to the coverage track that fastder sees. In *template + variant* (ASimulatoR default) both the canonical transcript and the alternative form are simulated, so an exon that the variant skips still receives reads from the template and shows up as a coverage dip rather than a hard zero. In *variant only* the FASTQ is post-filtered to keep reads sourced from the alternative transcript and the truth GFF is rewritten to remove templates, so the alternative-splicing signal is unambiguous: the skipped exon has zero coverage and the truth contains only the alternative isoform.

## Per-event Jaccard against best reference partner

For each reference transcript, the highest exonic Jaccard against any fastder transcript on the same strand. A value of 1 means the fastder match has the same exon set; 0 means no overlap.

```{r pub_jaccard, fig.width = 14, fig.height = 10}
if (!is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0) {
  fj <- fuzzy$jaccard %>%
    mutate(jaccard = as.numeric(jaccard),
           as_event = label_as(sample),
           scenario = relabel_scenario(scenario))
  ggplot(fj, aes(x = scenario, y = jaccard, fill = scenario)) +
    geom_violin(alpha = 0.4, scale = "width", trim = TRUE,
                colour = "grey40") +
    geom_jitter(width = 0.18, size = 0.25, alpha = 0.2, colour = "grey15") +
    stat_summary(fun = median, geom = "point", shape = 21, size = 2.6,
                 colour = "black", fill = "white", stroke = 0.7) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    labs(x = NULL, y = "Best exonic Jaccard against reference") +
    coord_cartesian(ylim = c(0, 1)) +
    facet_grid(param_id ~ as_event) +
    pub_theme +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          strip.text.y = element_text(size = 8))
} else {
  cat("no fuzzy_jaccard data")
}
```

## Per-event boundary precision

Distance from each fastder exon's start to the nearest reference exon start (left), and from end to nearest end (right). Same strand. Distances are clipped to ±200 bp for display.

Cumulative fraction of fastder boundaries that fall within an absolute distance to the nearest same-strand reference boundary. Each curve answers "what fraction of fastder's exon starts (or ends) are within X bp of a true splice site?". The x axis is log-scaled so the small-distance behaviour, where most boundaries live, is visible.

```{r pub_boundary, fig.width = 16, fig.height = 8}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances %>%
    mutate(distance = as.integer(distance),
           scenario = relabel_scenario(scenario),
           boundary = factor(boundary, levels = c("start", "end"),
                             labels = c("Exon start", "Exon end")),
           abs_distance = pmax(abs(distance), 1L))
  ggplot(fd, aes(x = abs_distance, colour = sample, linetype = scenario)) +
    stat_ecdf(geom = "step", linewidth = 0.7) +
    scale_colour_manual(values = as_event_palette, name = NULL,
                        labels = as_event_labels[as_event_levels]) +
    scale_linetype_manual(values = c("template + variant" = "solid",
                                     "variant only" = "32"),
                          name = NULL) +
    scale_x_log10(breaks = c(1, 5, 20, 100, 1000, 10000),
                  labels = c("1", "5", "20", "100", "1k", "10k")) +
    coord_cartesian(ylim = c(0, 1)) +
    facet_wrap(boundary ~ param_id, ncol = 5) +
    labs(x = "Absolute distance to nearest reference boundary (bp, log scale)",
         y = "Cumulative fraction of fastder boundaries") +
    pub_theme +
    theme(aspect.ratio = 0.7, legend.position = "bottom")
} else {
  cat("no fuzzy_distances data")
}
```

Companion summary: fraction of fastder boundaries within +/- 5 bp, +/- 20 bp, and +/- 50 bp of a same-strand reference boundary.

```{r pub_boundary_table, results = "asis"}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances %>%
    mutate(distance = as.integer(distance),
           boundary = factor(boundary, levels = c("start", "end"),
                             labels = c("Exon start", "Exon end")))
  bt <- fd %>%
    group_by(scenario, boundary, sample, param_id) %>%
    summarise(
      n = n(),
      pct_within_5 = mean(abs(distance) <= 5) * 100,
      pct_within_20 = mean(abs(distance) <= 20) * 100,
      pct_within_50 = mean(abs(distance) <= 50) * 100,
      median_abs = median(abs(distance)),
      .groups = "drop"
    )
  collapsible_kable(bt, "Show boundary-precision table",
                    digits = 2,
                    col.names = c("Scenario", "Boundary", "AS event", "Parameters", "n",
                                  "% within 5 bp", "% within 20 bp",
                                  "% within 50 bp", "Median |distance|"))
}
```

## Per-event strand concordance

```{r pub_strand, fig.width = 16, fig.height = 8}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs$category <- factor(fs$category, levels = c("concordant", "discordant", "unstranded", "unmatched"))
  fs$as_event <- label_as(fs$sample)
  fs$scenario <- relabel_scenario(fs$scenario)
  fs_pct <- fs %>%
    group_by(scenario, sample, as_event, param_id) %>%
    mutate(pct = n_fastder_transcripts / sum(n_fastder_transcripts) * 100) %>%
    ungroup()
  ggplot(fs_pct, aes(y = as_event, x = pct, fill = category)) +
    geom_col() +
    scale_fill_brewer(palette = "Set2", name = NULL) +
    labs(x = "Fraction of fastder transcripts (%)", y = NULL) +
    facet_grid(scenario ~ param_id) +
    pub_theme +
    theme(aspect.ratio = NULL, legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("no fuzzy_strand data")
}
```

## Per-event locus-recall curve

Fraction of reference loci with at least f of their exonic length covered by any predicted exon. Colour is the AS event class; linetype is the tool. Each baseline is shown at its `mc<v>` (and `pt0` for derfinder) matching the panel's fastder param_id `mc`/`pt`. fastder param sweep is faceted; baselines repeat across the columns of the same `mc`/`pt` slice.

This metric rewards broad coverage. Baselines emit one large region per coverage run that bleeds into intronic positions, which counts as coverage of any ref exon it intersects, so they sit high on this curve regardless of boundary accuracy. fastder emits short ER calls with tight boundaries and so trades locus recall for boundary precision (see Boundary precision and Multi-exon transcripts above).

```{r pub_recall, fig.width = 16, fig.height = 8}
if (!is.null(fuzzy_all$recall) && nrow(fuzzy_all$recall) > 0) {
  pids <- sort(unique(fuzzy_all$recall$param_id[fuzzy_all$recall$tool == "fastder"]))
  pid_join <- tibble::tibble(param_id_facet = pids,
                             mc = vapply(pids, extract_mc, FUN.VALUE = numeric(1)),
                             pt = vapply(pids, extract_pt, FUN.VALUE = numeric(1)))
  fr_fast <- fuzzy_all$recall %>%
    filter(tool == "fastder") %>%
    mutate(param_id_facet = param_id)
  fr_der <- fuzzy_all$recall %>%
    filter(tool == "derfinder") %>%
    mutate(mc = extract_mc(param_id), pt = extract_pt(param_id)) %>%
    inner_join(pid_join, by = c("mc", "pt"))
  fr_meg <- fuzzy_all$recall %>%
    filter(tool == "megadepth_baseline") %>%
    mutate(mc = extract_mc(param_id)) %>%
    inner_join(pid_join %>% select(param_id_facet, mc), by = "mc")
  fr <- bind_rows(fr_fast, fr_der, fr_meg) %>%
    mutate(threshold = as.numeric(threshold),
           recall = as.numeric(recall),
           scenario = relabel_scenario(scenario),
           tool = factor(tool, levels = c("megadepth_baseline",
                                          "derfinder", "fastder"))) %>%
    arrange(scenario, sample, param_id_facet, tool, threshold)
  ggplot(fr, aes(x = threshold, y = recall, colour = sample, linetype = tool,
                 group = interaction(scenario, sample, param_id_facet, tool))) +
    geom_line(linewidth = 0.7, alpha = 0.85) +
    scale_colour_manual(values = as_event_palette, name = NULL,
                        labels = as_event_labels[as_event_levels]) +
    scale_linetype_manual(values = c(megadepth_baseline = "11",
                                     derfinder = "32",
                                     fastder = "solid"), name = NULL) +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "Exonic coverage threshold",
         y = "Fraction of reference loci above threshold") +
    facet_wrap(scenario ~ param_id_facet, ncol = 5) +
    pub_theme +
    theme(legend.position = "bottom")
} else {
  cat("no fuzzy_recall data")
}
```

## Per-event scorecard

Headline numbers per AS event class.

```{r pub_scorecard, results = "asis"}
if (
  !is.null(fuzzy$jaccard) && !is.null(fuzzy$strand) && !is.null(fuzzy$recall) &&
  nrow(fuzzy$jaccard) > 0 && nrow(fuzzy$strand) > 0 && nrow(fuzzy$recall) > 0
) {
  jrow <- fuzzy$jaccard %>%
    mutate(jaccard = as.numeric(jaccard)) %>%
    group_by(scenario, sample, param_id) %>%
    summarise(median_jaccard = median(jaccard),
              pct_jaccard_ge_0_5 = mean(jaccard >= 0.5) * 100,
              .groups = "drop")
  srow <- fuzzy$strand %>%
    mutate(n = as.integer(n_fastder_transcripts)) %>%
    pivot_wider(id_cols = c(scenario, sample, param_id), names_from = category,
                values_from = n, values_fill = 0) %>%
    mutate(total = concordant + discordant + unstranded + unmatched,
           pct_concordant = concordant / total * 100,
           pct_discordant = discordant / total * 100) %>%
    select(scenario, sample, param_id, pct_concordant, pct_discordant)
  rrow <- fuzzy$recall %>%
    mutate(threshold = as.numeric(threshold), recall = as.numeric(recall)) %>%
    filter(threshold == 0.5) %>%
    select(scenario, sample, param_id, locus_recall_at_0_5 = recall) %>%
    mutate(locus_recall_at_0_5 = locus_recall_at_0_5 * 100)
  scorecard <- jrow %>%
    left_join(srow, by = c("scenario", "sample", "param_id")) %>%
    left_join(rrow, by = c("scenario", "sample", "param_id")) %>%
    mutate(as_event = label_as(sample)) %>%
    arrange(scenario, param_id, sample) %>%
    select(scenario, param_id, as_event, median_jaccard, pct_jaccard_ge_0_5,
           pct_concordant, pct_discordant, locus_recall_at_0_5)
  collapsible_kable(scorecard, "Show per-event scorecard",
                    digits = 2,
                    col.names = c("Scenario", "Parameters", "AS event", "Median Jaccard",
                                  "% Jaccard >= 0.5", "% strand concordant",
                                  "% strand discordant", "% loci with >= 50% coverage"))
}
```

## Parameter sensitivity

Three top-line metrics (median Jaccard, % strand concordant, locus recall at coverage threshold 0.5) are aggregated across samples and plotted against each fastder parameter axis individually. The other parameter axes appear as separate jittered points in each panel, so a parameter that controls the result will show a clean monotonic trend, while a parameter the result is insensitive to will show a flat cloud. Lines connect the means at each parameter value to make the trend visible.

```{r param_sensitivity, fig.width = 7, fig.height = 6}
have_data <- !is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0 &&
             !is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0 &&
             !is.null(fuzzy$recall) && nrow(fuzzy$recall) > 0

if (have_data) {
  jagg <- fuzzy$jaccard %>%
    parse_param_id() %>%
    mutate(jaccard = as.numeric(jaccard)) %>%
    group_by(scenario, param_id, min_coverage, min_length, coverage_tolerance, position_tolerance) %>%
    summarise(median_jaccard = median(jaccard), .groups = "drop")

  sagg <- fuzzy$strand %>%
    parse_param_id() %>%
    mutate(n = as.integer(n_fastder_transcripts)) %>%
    pivot_wider(id_cols = c(scenario, param_id, min_coverage, min_length,
                            coverage_tolerance, position_tolerance, sample),
                names_from = category, values_from = n, values_fill = 0) %>%
    mutate(total = concordant + discordant + unstranded + unmatched,
           pct_concordant = ifelse(total > 0, concordant / total * 100, NA_real_)) %>%
    group_by(scenario, param_id, min_coverage, min_length,
             coverage_tolerance, position_tolerance) %>%
    summarise(pct_concordant = mean(pct_concordant, na.rm = TRUE), .groups = "drop")

  ragg <- fuzzy$recall %>%
    parse_param_id() %>%
    mutate(threshold = as.numeric(threshold), recall = as.numeric(recall)) %>%
    filter(threshold == 0.5) %>%
    group_by(scenario, param_id, min_coverage, min_length,
             coverage_tolerance, position_tolerance) %>%
    summarise(locus_recall_at_0_5 = mean(recall) * 100, .groups = "drop")

  combined <- jagg %>%
    left_join(sagg, by = c("scenario", "param_id", "min_coverage", "min_length",
                           "coverage_tolerance", "position_tolerance")) %>%
    left_join(ragg, by = c("scenario", "param_id", "min_coverage", "min_length",
                           "coverage_tolerance", "position_tolerance")) %>%
    mutate(scenario = relabel_scenario(scenario))

  param_long <- combined %>%
    pivot_longer(c(min_coverage, min_length, coverage_tolerance, position_tolerance),
                 names_to = "param", values_to = "param_value") %>%
    filter(!is.na(param_value)) %>%
    mutate(param = recode(param,
                          min_coverage = "min_coverage",
                          min_length = "min_length",
                          coverage_tolerance = "coverage_tolerance",
                          position_tolerance = "position_tolerance"))

  metric_long <- param_long %>%
    pivot_longer(c(median_jaccard, pct_concordant, locus_recall_at_0_5),
                 names_to = "metric", values_to = "metric_value") %>%
    mutate(metric = recode(metric,
                           median_jaccard = "Median Jaccard",
                           pct_concordant = "% strand concordant",
                           locus_recall_at_0_5 = "% loci with >= 50% coverage"))

  ggplot(metric_long, aes(x = param_value, y = metric_value,
                          colour = scenario, group = scenario)) +
    geom_jitter(size = 0.8, width = 0.02, alpha = 0.4) +
    stat_summary(fun = mean, geom = "line", linewidth = 0.7) +
    stat_summary(fun = mean, geom = "point", size = 1.8) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    facet_grid(metric ~ param, scales = "free") +
    labs(x = "Parameter value", y = "Metric value") +
    pub_theme +
    theme(legend.position = "bottom",
          aspect.ratio = 1,
          axis.text = element_text(size = 7),
          strip.text = element_text(size = 7))
}
```

Reading the figure: rows are the three top-line metrics, columns are the four fastder parameter axes (axes that are not in the current grid show no points). Within a panel, points share the same value for the column's parameter; the spread within a value comes from the other parameter axes. A steep slope across the parameter values means the result is sensitive to that parameter; a flat scatter means it is not.

# gffcompare metrics

`r summary_path`

```{r summary_table, results = "asis"}
collapsible_kable(df, "Show full gffcompare summary table", digits = 2)
```

## Sensitivity and precision across all levels

Every row in `summary.csv` carries sensitivity and precision at the six gffcompare levels. The plot below shows them on the same axes per scenario. Each coloured point is one (sample, parameter combination) at one gffcompare level. For every (scenario, level) cell the best-performing (sample, parameter) combination, defined as the one with the highest mean of sensitivity and precision, is circled in black and labelled with its AS event sample and parameter id, so the parameter combination that maximises each level is visible at a glance.

```{r levels_scatter, fig.width = 6, fig.height = 4}
levels_long <- df %>%
  pivot_longer(
    cols = matches("_(sens|prec)$"),
    names_to = c("level", "metric"),
    names_pattern = "(.+)_(sens|prec)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  mutate(scenario = relabel_scenario(scenario))

level_label <- function(x) {
  recode(x,
    base = "Base",
    exon = "Exon",
    intron = "Intron",
    intron_chain = "Intron chain",
    transcript = "Transcript",
    locus = "Locus"
  )
}
levels_long$level <- level_label(levels_long$level)

levels_long <- levels_long %>%
  mutate(combined = (sens + prec) / 2)

best_per_facet <- levels_long %>%
  group_by(scenario, level) %>%
  slice_max(combined, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(label = paste(sample, param_id, sep = "\n"))

ggplot(levels_long, aes(x = sens, y = prec, colour = level, shape = level)) +
  geom_point(size = 2.2, alpha = 0.6) +
  geom_point(data = best_per_facet, size = 3, stroke = 1.2,
             colour = "black", fill = NA, shape = 21) +
  geom_text_repel(data = best_per_facet,
                  aes(label = label, colour = level),
                  size = 2.6, lineheight = 0.85,
                  box.padding = 0.4, max.overlaps = 20,
                  show.legend = FALSE) +
  scale_shape_manual(values = c(16, 17, 15, 18, 4, 3)) +
  coord_fixed(xlim = c(0, 100), ylim = c(0, 100)) +
  facet_wrap(~ scenario, ncol = 2) +
  labs(x = "Sensitivity (%)", y = "Precision (%)", colour = "Level", shape = "Level") +
  pub_theme +
  theme(legend.position = "bottom")
```

### All tools, stratified by level and scenario

Same six gffcompare levels but with all three tools overlaid. One sub-panel per (level, scenario), tool on the x-axis. Both sensitivity and precision are shown.

```{r levels_scatter_per_tool, fig.width = 12, fig.height = 7}
if ("tool" %in% names(df_all) && nrow(df_all) > 0) {
  levels_all <- df_all %>%
    pivot_longer(
      cols = matches("_(sens|prec)$"),
      names_to = c("level", "metric"),
      names_pattern = "(.+)_(sens|prec)",
      values_to = "value"
    ) %>%
    mutate(scenario = relabel_scenario(scenario),
           level = level_label(level),
           metric = recode(metric, sens = "Sensitivity", prec = "Precision"),
           value = as.numeric(value))
  ggplot(levels_all, aes(x = tool, y = value, colour = tool, shape = tool)) +
    geom_jitter(width = 0.15, height = 0, size = 1.6, alpha = 0.4) +
    stat_summary(fun = median, geom = "point", shape = 21, size = 2.6,
                 colour = "black", fill = "white", stroke = 0.7) +
    scale_colour_brewer(palette = "Set2", name = NULL) +
    scale_shape_manual(values = c(fastder = 16, derfinder = 17,
                                  megadepth_baseline = 15), name = NULL) +
    coord_cartesian(ylim = c(0, 100)) +
    facet_grid(metric ~ level + scenario,
               labeller = labeller(.cols = label_value)) +
    labs(x = NULL, y = "Value (%)") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL,
          legend.position = "bottom")
}
```

Each point is one (sample, param_id) cell; the hollow black point marks the median per tool in that (level, scenario, metric) facet. fastder leads on transcript / intron_chain / locus when there is signal at all (see Note above on metric choice).

## Per-level bars

```{r level_bars, fig.width = 6, fig.height = 4}
levels_long_metric <- df %>%
  pivot_longer(
    cols = matches("_(sens|prec)$"),
    names_to = c("level", "metric"),
    names_pattern = "(.+)_(sens|prec)",
    values_to = "value"
  ) %>%
  mutate(scenario = relabel_scenario(scenario),
         level = level_label(level),
         metric = recode(metric, sens = "Sensitivity", prec = "Precision"))

ggplot(levels_long_metric, aes(x = level, y = value, fill = scenario)) +
  geom_boxplot(alpha = 0.6, outlier.size = 0.5) +
  facet_wrap(~ metric, ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Level", y = "Value (%)", fill = NULL) +
  pub_theme +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom", aspect.ratio = NULL)
```

Each box aggregates across samples and parameter combinations for one (level, scenario). The full per-(sample, param) breakdown is below as a heatmap.

### Per-(sample, param) heatmap

Top 20 (sample, parameter) rows for each of Sensitivity and Precision, ranked by mean across the six gffcompare levels. Cell numbers are percentages. The full table is in summary.csv.

```{r level_bars_heatmap, fig.width = 11, fig.height = 6}
TOP_N_HEATMAP_ROWS <- 20
# colorRamp2 fails if start == end, which happens whenever a parameter axis
# has only one distinct value in the swept grid (e.g. config_quick uses a
# single min_coverage). Pad to a flat single-color ramp in that case so the
# annotation still renders.
safe_ramp <- function(values, high) {
  vals <- suppressWarnings(as.numeric(values))
  vals <- vals[is.finite(vals)]
  if (length(unique(vals)) < 2) {
    v <- if (length(vals) == 0) 0 else vals[1]
    return(colorRamp2(c(v - 1, v + 1), c(high, high)))
  }
  colorRamp2(range(vals), c("white", high))
}

build_level_heatmap <- function(metric_name, palette_high) {
  d <- levels_long_metric %>%
    filter(metric == metric_name) %>%
    parse_param_id()
  mat <- d %>%
    mutate(row_id = paste(scenario, sample, param_id, sep = " | ")) %>%
    pivot_wider(id_cols = row_id, names_from = level, values_from = value)
  m <- as.matrix(mat[, -1])
  rownames(m) <- mat$row_id
  # Sort rows by mean across levels, best on top, then trim to the top-N to
  # keep the heatmap legible for full parameter sweeps.
  row_order <- order(rowMeans(m, na.rm = TRUE), decreasing = TRUE)
  m <- m[row_order, , drop = FALSE]
  if (nrow(m) > TOP_N_HEATMAP_ROWS) {
    m <- m[seq_len(TOP_N_HEATMAP_ROWS), , drop = FALSE]
  }
  meta <- d %>%
    mutate(row_id = paste(scenario, sample, param_id, sep = " | ")) %>%
    distinct(row_id, scenario, sample, min_coverage, min_length, coverage_tolerance) %>%
    arrange(match(row_id, rownames(m))) %>%
    filter(row_id %in% rownames(m))
  ra <- rowAnnotation(
    Scenario = as.character(meta$scenario),
    `AS event` = meta$sample,
    `min_coverage` = meta$min_coverage,
    `min_length` = meta$min_length,
    `coverage_tol` = meta$coverage_tolerance,
    col = list(
      Scenario = c("template + variant" = "#66c2a5", "variant only" = "#fc8d62"),
      `AS event` = as_event_palette,
      `min_coverage` = safe_ramp(meta$min_coverage, "#08306b"),
      `min_length` = safe_ramp(meta$min_length, "#7f2704"),
      `coverage_tol` = safe_ramp(meta$coverage_tolerance, "#3f007d")
    ),
    annotation_name_side = "top",
    annotation_name_gp = gpar(fontsize = 8),
    show_legend = TRUE
  )
  Heatmap(m, name = metric_name,
          col = colorRamp2(c(0, 50, 100), c("white", "#fdb462", palette_high)),
          left_annotation = ra,
          cluster_rows = FALSE, cluster_columns = FALSE,
          row_names_side = "left",
          row_names_gp = gpar(fontsize = 6),
          column_names_gp = gpar(fontsize = 8),
          column_names_rot = 30,
          width = unit(7, "cm"),
          show_row_names = FALSE,
          cell_fun = function(j, i, x, y, width, height, fill) {
            grid.text(sprintf("%.0f", m[i, j]), x, y,
                      gp = gpar(fontsize = 5.5,
                                col = ifelse(m[i, j] > 60, "white", "grey15")))
          })
}
ht_sens <- build_level_heatmap("Sensitivity", "#1b6db0")
ht_prec <- build_level_heatmap("Precision", "#1b6db0")
draw(ht_sens + ht_prec, ht_gap = unit(2, "mm"),
     heatmap_legend_side = "bottom",
     annotation_legend_side = "bottom",
     merge_legend = TRUE)
```

## Missed and novel features

In gffcompare's terminology, "missed" means a feature exists in the reference but was not recovered by fastder (a false negative against the truth). "Novel" means fastder reported a feature that does not appear in the reference (a false positive). Because the reference here is the simulated splicing_variants.gff3, novel features were *not* simulated and so are by construction false positives. Most of them come from coverage drop-offs that shifted the boundary just enough that gffcompare doesn't accept the match, plus a small number of stray STAR alignments off-target.

```{r missed_novel, fig.width = 6, fig.height = 4}
mn <- df %>%
  pivot_longer(
    cols = matches("_(pct|count|total)$"),
    names_to = c("kind", "feature", "field"),
    names_pattern = "(missed|novel)_(exons|introns|loci)_(pct|count|total)",
    values_to = "value"
  ) %>%
  filter(field == "pct") %>%
  mutate(scenario = relabel_scenario(scenario),
         feature = recode(feature, exons = "Exons", introns = "Introns", loci = "Loci"),
         kind = recode(kind, missed = "Missed (in reference, not recovered)",
                             novel = "Novel (in fastder, not in reference)"))

ggplot(mn, aes(x = feature, y = value, fill = scenario)) +
  geom_boxplot(alpha = 0.6, outlier.size = 0.5) +
  facet_wrap(~ kind, ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Fraction (%)", fill = NULL) +
  pub_theme +
  theme(legend.position = "bottom", aspect.ratio = NULL)
```

Each box aggregates across samples and parameter combinations for one (feature, scenario). The full per-(sample, param) breakdown is below as a heatmap.

### Per-(sample, param) missed/novel heatmap

Top 20 (sample, parameter) rows by lowest mean missed or novel rate (best on top).

```{r missed_novel_heatmap, fig.width = 10, fig.height = 6}
build_mn_heatmap <- function(kind_name, palette_high) {
  d <- mn %>%
    filter(kind == kind_name) %>%
    parse_param_id()
  mat <- d %>%
    mutate(row_id = paste(scenario, sample, param_id, sep = " | ")) %>%
    pivot_wider(id_cols = row_id, names_from = feature, values_from = value)
  m <- as.matrix(mat[, -1])
  rownames(m) <- mat$row_id
  row_order <- order(rowMeans(m, na.rm = TRUE), decreasing = FALSE)
  m <- m[row_order, , drop = FALSE]
  if (nrow(m) > TOP_N_HEATMAP_ROWS) {
    m <- m[seq_len(TOP_N_HEATMAP_ROWS), , drop = FALSE]
  }
  meta <- d %>%
    mutate(row_id = paste(scenario, sample, param_id, sep = " | ")) %>%
    distinct(row_id, scenario, sample, min_coverage, min_length, coverage_tolerance) %>%
    arrange(match(row_id, rownames(m))) %>%
    filter(row_id %in% rownames(m))
  ra <- rowAnnotation(
    Scenario = as.character(meta$scenario),
    `AS event` = meta$sample,
    `min_coverage` = meta$min_coverage,
    `min_length` = meta$min_length,
    `coverage_tol` = meta$coverage_tolerance,
    col = list(
      Scenario = c("template + variant" = "#66c2a5", "variant only" = "#fc8d62"),
      `AS event` = as_event_palette,
      `min_coverage` = safe_ramp(meta$min_coverage, "#08306b"),
      `min_length` = safe_ramp(meta$min_length, "#7f2704"),
      `coverage_tol` = safe_ramp(meta$coverage_tolerance, "#3f007d")
    ),
    annotation_name_side = "top",
    annotation_name_gp = gpar(fontsize = 8)
  )
  Heatmap(m, name = kind_name,
          col = colorRamp2(c(0, 50, 100), c("white", "#fcae91", palette_high)),
          left_annotation = ra,
          cluster_rows = FALSE, cluster_columns = FALSE,
          column_names_gp = gpar(fontsize = 8),
          column_names_rot = 30,
          width = unit(3.5, "cm"),
          show_row_names = FALSE,
          cell_fun = function(j, i, x, y, width, height, fill) {
            grid.text(sprintf("%.1f", m[i, j]), x, y,
                      gp = gpar(fontsize = 5.5,
                                col = ifelse(m[i, j] > 60, "white", "grey15")))
          })
}
ht_missed <- build_mn_heatmap("Missed (in reference, not recovered)", "#67000d")
ht_novel <- build_mn_heatmap("Novel (in fastder, not in reference)", "#67000d")
draw(ht_missed + ht_novel, ht_gap = unit(2, "mm"),
     heatmap_legend_side = "bottom",
     annotation_legend_side = "bottom",
     merge_legend = TRUE)
```

## Matching counts

```{r matching_counts, results = "asis"}
m_cols <- grep("^matching_", names(df), value = TRUE)
if (length(m_cols) > 0) {
  m <- df %>% select(scenario, sample, param_id, all_of(m_cols))
  collapsible_kable(m, "Show matching counts", digits = 0)
}
```

## Query and reference scope

```{r scope_table, results = "asis"}
scope_cols <- intersect(
  c("query_mrnas", "query_loci", "query_multi_exon",
    "ref_mrnas", "ref_loci", "ref_multi_exon", "super_loci"),
  names(df)
)
if (length(scope_cols) > 0) {
  collapsible_kable(df %>% select(scenario, sample, param_id, all_of(scope_cols)),
                    "Show query and reference scope", digits = 0)
}
```

# Stitched chain stats

`r chain_path` (`r nrow(cs)` transcripts across `r length(unique(cs$param_id))` parameter combinations)

The score column reflects the per-exon mean coverage written by fastder into GTF column 6, which is what drives the "mean coverage" axis below. The truth set (the simulated splicing_variants.gff3) is overlaid in red where available, so the rendered output can be compared directly with what was simulated. The Parameters column lists each fastder parameter combination, abbreviated; mc0.01 stands for min_coverage = 0.01. The pipeline's parameter grid has one combination here because `fastder.min_coverage` in the config is a one-element list; adding more values, or other knobs like min_length, position_tolerance, or coverage_tolerance, would expand the grid and produce more rows in this table.

```{r chain_summary_table, results = "asis"}
chain_summary <- cs %>%
  group_by(scenario, param_id) %>%
  summarise(
    n_transcripts = n(),
    pct_singletons = mean(n_exons == 1) * 100,
    median_exons = median(n_exons),
    mean_exons = mean(n_exons),
    max_exons = max(n_exons),
    median_length_bp = median(total_exon_length),
    median_score = median(score),
    .groups = "drop"
  )
collapsible_kable(chain_summary, "Show chain summary",
                  digits = 2,
                  col.names = c("Scenario", "Parameters", "Transcripts",
                                "% single-exon", "Median exons", "Mean exons",
                                "Max exons", "Median exonic length (bp)",
                                "Median mean coverage"))
```

## Number of exons per transcript

The grey bars show fastder's per-transcript exon count distribution. The red step trace is the truth distribution computed from the simulated splicing_variants.gff3. A close match means fastder is recovering the correct number of pieces per gene. The x axis is log10-scaled because most transcripts are single-exon and the long tail (one or two transcripts with many exons) compresses everything else otherwise.

```{r n_exons_hist, fig.width = 12, fig.height = 6}
cs_plot <- cs %>% mutate(scenario = relabel_scenario(scenario))
truth_plot <- if (!is.null(truth_chain)) {
  truth_chain %>%
    mutate(scenario = relabel_scenario(scenario)) %>%
    count(scenario, n_exons)
} else NULL

p <- ggplot(cs_plot, aes(x = n_exons)) +
  geom_histogram(fill = "grey60", colour = "grey30",
                 breaks = c(0.5, 1.5, 2.5, 3.5, 4.5, 5.5,
                            7.5, 10.5, 15.5, 20.5, 30.5, 50.5, 100.5, 1000)) +
  facet_grid(scenario ~ param_id) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100)) +
  labs(x = "Number of exons per transcript (log scale)",
       y = "Number of transcripts") +
  pub_theme
if (!is.null(truth_plot) && nrow(truth_plot) > 0) {
  p <- p + geom_step(data = truth_plot,
                     aes(x = n_exons, y = n),
                     direction = "vh", colour = "firebrick", linewidth = 0.7)
}
p
```

## Per-chromosome transcript count

```{r per_chrom_count, fig.width = 14, fig.height = 5}
chrom_counts <- cs %>%
  count(scenario, param_id, chrom) %>%
  mutate(scenario = relabel_scenario(scenario))
truth_chrom <- if (!is.null(truth_chain)) {
  # truth has no param_id; we replicate it across parameter columns by joining
  truth_chain %>%
    mutate(scenario = relabel_scenario(scenario)) %>%
    count(scenario, chrom) %>%
    rename(n_truth = n)
} else NULL

p <- ggplot(chrom_counts, aes(x = chrom, y = n)) +
  geom_col(fill = "grey60", colour = "grey30") +
  facet_grid(scenario ~ param_id) +
  coord_flip() +
  labs(x = "Chromosome", y = "Number of transcripts") +
  pub_theme +
  theme(aspect.ratio = NULL)
if (!is.null(truth_chrom) && nrow(truth_chrom) > 0) {
  p <- p + geom_point(data = truth_chrom,
                      aes(x = chrom, y = n_truth),
                      colour = "firebrick", size = 1.5)
}
p
```

## Total exonic length per transcript

```{r length_hist, fig.width = 14, fig.height = 6}
cs_plot <- cs %>% mutate(scenario = relabel_scenario(scenario))
ggplot(cs_plot, aes(x = total_exon_length)) +
  geom_histogram(bins = 30, fill = "grey60", colour = "grey30") +
  facet_grid(scenario ~ param_id) +
  scale_x_log10() +
  labs(x = "Total exonic length per transcript (bp)",
       y = "Number of transcripts") +
  pub_theme
```

## Mean coverage vs number of exons

```{r score_vs_n_exons, fig.width = 14, fig.height = 6}
cs_plot <- cs %>% mutate(scenario = relabel_scenario(scenario))
ggplot(cs_plot, aes(x = n_exons, y = score)) +
  geom_point(alpha = 0.3) +
  scale_y_log10() +
  facet_grid(scenario ~ param_id) +
  labs(x = "Number of exons per transcript", y = "Mean coverage") +
  pub_theme
```

## Mean coverage vs total exonic length

```{r score_vs_length, fig.width = 14, fig.height = 6}
cs_plot <- cs %>% mutate(scenario = relabel_scenario(scenario))
ggplot(cs_plot, aes(x = total_exon_length, y = score)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  facet_grid(scenario ~ param_id) +
  labs(x = "Total exonic length per transcript (bp)",
       y = "Mean coverage") +
  pub_theme
```

# Fuzzy fastder vs reference

These metrics drop gffcompare's exact-boundary requirement and look at how much fastder transcripts overlap reference transcripts on the same strand. Useful for fastder because its exon boundaries come from coverage drop-offs, not splice-site coordinates, so it never wins on transcript-exact matching.

## Reciprocal-best Jaccard per reference transcript

For each reference transcript, the highest exonic Jaccard against any fastder transcript on the same strand. A value of 1 means perfect exonic match; 0 means no overlap.

```{r jaccard_hist, fig.width = 6, fig.height = 6}
if (!is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0) {
  fj <- fuzzy$jaccard
  fj$jaccard <- as.numeric(fj$jaccard)
  ggplot(fj, aes(x = jaccard)) +
    geom_histogram(bins = 30, boundary = 0) +
    facet_grid(sample ~ param_id) +
    theme_bw() +
    theme(aspect.ratio = 1, legend.position = "none")
} else {
  cat("no fuzzy_jaccard data")
}
```

```{r jaccard_summary, results = "asis"}
if (!is.null(fuzzy$jaccard) && nrow(fuzzy$jaccard) > 0) {
  fj <- fuzzy$jaccard
  fj$jaccard <- as.numeric(fj$jaccard)
  jsum <- fj %>%
    group_by(scenario, sample, param_id) %>%
    summarise(
      n_ref_transcripts = n(),
      pct_jaccard_zero = mean(jaccard == 0) * 100,
      median_jaccard = median(jaccard),
      mean_jaccard = mean(jaccard),
      pct_jaccard_ge_0_5 = mean(jaccard >= 0.5) * 100,
      pct_jaccard_ge_0_9 = mean(jaccard >= 0.9) * 100,
      .groups = "drop"
    )
  collapsible_kable(jsum, "Show Jaccard summary", digits = 2)
}
```

## Boundary-distance ECDF

Cumulative fraction of fastder boundaries within an absolute distance to the nearest same-strand reference boundary. Same panel layout as the publication figure: rows are exon start / exon end, columns are scenarios. Colours highlight the AS event class.

```{r boundary_hist, fig.width = 16, fig.height = 8}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances %>%
    mutate(distance = as.integer(distance),
           scenario = relabel_scenario(scenario),
           boundary = factor(boundary, levels = c("start", "end"),
                             labels = c("Exon start", "Exon end")),
           abs_distance = pmax(abs(distance), 1L))
  ggplot(fd, aes(x = abs_distance, colour = sample, linetype = scenario)) +
    stat_ecdf(geom = "step", linewidth = 0.7) +
    scale_colour_manual(values = as_event_palette, name = NULL,
                        labels = as_event_labels[as_event_levels]) +
    scale_linetype_manual(values = c("template + variant" = "solid",
                                     "variant only" = "32"),
                          name = NULL) +
    scale_x_log10(breaks = c(1, 5, 20, 100, 1000, 10000),
                  labels = c("1", "5", "20", "100", "1k", "10k")) +
    coord_cartesian(ylim = c(0, 1)) +
    facet_wrap(boundary ~ param_id, ncol = 5) +
    labs(x = "Absolute distance to nearest reference boundary (bp, log scale)",
         y = "Cumulative fraction of fastder boundaries") +
    pub_theme +
    theme(aspect.ratio = 0.7, legend.position = "bottom")
} else {
  cat("no fuzzy_distances data")
}
```

```{r boundary_summary, results = "asis"}
if (!is.null(fuzzy$distances) && nrow(fuzzy$distances) > 0) {
  fd <- fuzzy$distances
  fd$distance <- as.integer(fd$distance)
  bsum <- fd %>%
    group_by(scenario, sample, param_id, boundary) %>%
    summarise(
      n = n(),
      pct_within_5bp = mean(abs(distance) <= 5) * 100,
      pct_within_50bp = mean(abs(distance) <= 50) * 100,
      median_abs = median(abs(distance)),
      .groups = "drop"
    )
  collapsible_kable(bsum, "Show boundary summary", digits = 2)
}
```

## Locus recall at fixed coverage thresholds

Fraction of reference loci with at least f of their exonic length covered by any fastder exon on the same strand, at thresholds f from 0.05 to 1.0 in steps of 0.05.

The two scenario curves often cross in a single panel: template + variant tends to win at low thresholds (extra reads from both isoforms hit more loci even if only partially), while variant only wins at high thresholds (the reads land more concentrated on the alternative form, so the loci that are covered are covered nearly fully). This is the classic depth-versus-breadth trade-off, not an artefact.

```{r locus_recall_curve, fig.width = 8, fig.height = 12}
if (!is.null(fuzzy$recall) && nrow(fuzzy$recall) > 0) {
  fr <- fuzzy$recall %>%
    mutate(threshold = as.numeric(threshold),
           recall = as.numeric(recall),
           scenario = relabel_scenario(scenario)) %>%
    arrange(sample, param_id, scenario, threshold)
  ggplot(fr, aes(x = threshold, y = recall, colour = scenario,
                 group = interaction(scenario, sample, param_id))) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.6) +
    coord_cartesian(ylim = c(0, 1)) +
    facet_grid(sample ~ param_id) +
    labs(x = "Exonic coverage threshold",
         y = "Fraction of reference loci above threshold",
         colour = NULL) +
    pub_theme +
    theme(legend.position = "bottom")
} else {
  cat("no fuzzy_recall data")
}
```

```{r locus_recall_table, results = "asis"}
if (!is.null(fuzzy$recall) && nrow(fuzzy$recall) > 0) {
  collapsible_kable(fuzzy$recall, "Show full recall table", digits = 3)
}
```

## Strand concordance per fastder transcript

For each fastder StitchedER, the best-overlapping reference transcript on either strand is selected. The fastder strand is then classified as concordant (matches the partner), discordant (opposite of the partner), unstranded (fastder emitted '.'), or unmatched (no reference overlap). Counts are per scenario, sample, and fastder parameter combination. Parameter combinations are abbreviated, for instance mc0.01 means min_coverage = 0.01; the panel grid below has as many columns as parameter combinations declared in the config.

```{r strand_counts, fig.width = 16, fig.height = 8}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs$category <- factor(fs$category, levels = c("concordant", "discordant", "unstranded", "unmatched"))
  fs$scenario <- relabel_scenario(fs$scenario)
  use_param_facet <- length(unique(fs$param_id)) > 1
  p <- ggplot(fs, aes(x = sample, y = n_fastder_transcripts, fill = category)) +
    geom_col() +
    labs(x = NULL, y = "Number of fastder transcripts", fill = NULL) +
    pub_theme +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))
  if (use_param_facet) {
    p + facet_grid(scenario ~ param_id)
  } else {
    p + facet_wrap(~ scenario)
  }
} else {
  cat("no fuzzy_strand data")
}
```

```{r strand_pct, fig.width = 16, fig.height = 8}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs$category <- factor(fs$category, levels = c("concordant", "discordant", "unstranded", "unmatched"))
  fs$scenario <- relabel_scenario(fs$scenario)
  fs_pct <- fs %>%
    group_by(scenario, sample, param_id) %>%
    mutate(pct = n_fastder_transcripts / sum(n_fastder_transcripts) * 100) %>%
    ungroup()
  use_param_facet <- length(unique(fs_pct$param_id)) > 1
  p <- ggplot(fs_pct, aes(x = sample, y = pct, fill = category)) +
    geom_col() +
    labs(x = NULL, y = "Fraction of fastder transcripts (%)", fill = NULL) +
    pub_theme +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))
  if (use_param_facet) {
    p + facet_grid(scenario ~ param_id)
  } else {
    p + facet_wrap(~ scenario)
  }
}
```

```{r strand_table, results = "asis"}
if (!is.null(fuzzy$strand) && nrow(fuzzy$strand) > 0) {
  fs <- fuzzy$strand
  fs$n_fastder_transcripts <- as.integer(fs$n_fastder_transcripts)
  fs_wide <- fs %>%
    pivot_wider(names_from = category, values_from = n_fastder_transcripts, values_fill = 0) %>%
    mutate(total = concordant + discordant + unstranded + unmatched,
           pct_concordant = concordant / total * 100,
           pct_discordant = discordant / total * 100,
           pct_unstranded = unstranded / total * 100,
           pct_unmatched = unmatched / total * 100)
  collapsible_kable(fs_wide, "Show strand counts and fractions", digits = 2)
}
```
